---
title: "Cytokine-driven T-cell differentiation timecourse"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{ben_legg_cytokine_timecourse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteBuild{true}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

### 1. Workplace setup

```{r setup, warning = FALSE, include = FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(purrr)
library(forcats)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(edgeR)
library(org.Mm.eg.db)
library(msigdbr)
library(clusterProfiler)
library(pheatmap)
library(DT)
library(patchwork)
```

#### 1.1 Directory structure

```{r directory_setup, warning = FALSE, echo=F}
# 1. make sure the bioinf_scratch drive is mounted
# 2. session -> set working directory -> scripts

######## directory structure #######
projectDir=paste0("../")
resultsDir=paste0(projectDir,"/results")
imageDir=paste0(resultsDir,"/figures/")
robjectsDir = paste(resultsDir,"/Robjects/",sep="")
tableDir = paste(resultsDir,"/tables/",sep="")

system(paste("mkdir -p",robjectsDir))
system(paste("mkdir -p",tableDir))
system(paste("mkdir -p",imageDir))

# counts come from 5.feature_counts.sbatch, one file per sample
inDir<-paste0(resultsDir,"/counts/featurecounts/")
```

#### 1.2 Experimental design

Naive T-cells cultured with one of four cytokines across a timecourse, against
a negative control (NC) and an uncultured naive baseline.

1. **NC only exists at 24h, 48h and 72h.** The 96h and 120h cytokine samples
   have no time-matched control. 
2. **IL36b_96h is entirely batch 2.** All three of its replicates were
   resequenced in the `GMP5RJ` batch, along with IL2_96h rep 3. 



```{r colours, warning = FALSE, echo=F}
# IL2 / IL7_15 / IL36 keep the same colours used in the dane_rnaseq analysis so
# the two Ben projects can be read side by side. IL18 is new here and takes
# orange. Naive and NC are greys, because they are references rather than
# treatments of interest.
treatment_colours <- c(
  "Naive" = "gray70",
  "NC"    = "gray40",
  "IL2"   = "#F20D0D",
  "IL715" = "#00B050",
  "IL18"  = "#FF8C00",
  "IL36b" = "#0096FF"
)

# Direction colours, shared by the volcanoes and the pathway dotplot.
direction_colours <- c(
  "DOWN" = "#0096FF",
  "NO"   = "gray60",
  "UP"   = "#F20D0D"
)
```

```{r volcano_function, include = FALSE, echo=F}
# Same volcano used in 8.ben_volcanoes.Rmd. Pass highlight_genes to label a
# fixed marker list instead of every significant gene.
plot_volcano <- function(top,
                         FDR_cutoff = 0.05,
                         max.overlaps = 30,
                         highlight_genes = NULL) {
  top <- as.data.frame(top)

  # Rename column if needed
  if ("adj.P.Val" %in% colnames(top)) {
    top <- top %>%
      rename(FDR = adj.P.Val)
  }

  # Differential expression flags. Built with ifelse rather than by indexing
  # into a subset, so a contrast with no significant genes does not error.
  top$diff_expression <- "NO"
  top$diff_expression[top$logFC >= 1  & top$FDR <= FDR_cutoff] <- "UP"
  top$diff_expression[top$logFC <= -1 & top$FDR <= FDR_cutoff] <- "DOWN"

  # Label logic
  top$labelgenes <- ""
  if (!is.null(highlight_genes)) {
    highlight_genes_present <- intersect(highlight_genes, rownames(top))
    top[rownames(top) %in% highlight_genes_present, "labelgenes"] <-
      rownames(top)[rownames(top) %in% highlight_genes_present]
  } else {
    top[top$diff_expression != "NO", "labelgenes"] <-
      rownames(top[top$diff_expression != "NO", ])
  }

  ggplot(top, aes(x = logFC, y = -log10(FDR), color = diff_expression, label = labelgenes)) +
    geom_point(size = 1.8) +
    geom_text_repel(
      max.overlaps = max.overlaps,
      min.segment.length = 0,       # always draw line
      segment.color = "grey40",     # color of connecting line
      segment.size = 0.6,           # thickness of the line
      box.padding = 0.3,
      point.padding = 0.25,
      seed = 123                    # reproducible layout
    ) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = -log10(FDR_cutoff), linetype = "dashed", color = "darkred") +
    scale_color_manual(values = direction_colours) +
    theme_classic(base_size = 12) +
    labs(
      x = "log2 Fold Change",
      y = "-log10(FDR)",
      color = "Differential expression",
      title = "Volcano plot"
    )
}
```

### 2. QC

#### 2.1 Import data

```{r data_import, warning = FALSE, echo=F}
rObjectFile<-paste0(robjectsDir,"featurecounts_files.Rdata")

# featureCounts writes a "# Program:.." banner, then a wide table whose LAST
# column holds the counts and is named after the input BAM path.
if(!file.exists(rObjectFile)){
  inFiles<-list.files(inDir,full.names=T, pattern=".featureCounts.txt$")
  results<-list()
  for(inFile in inFiles){

    sampleName<-basename(inFile)
    sampleName<-gsub(".featureCounts.txt","",sampleName)
    sampleName<-gsub("_trimmed","",sampleName)

    data<-read.delim(inFile,comment.char="#")

    cat(sampleName)
    cat("\n")
    results[[sampleName]]<-data[,ncol(data)]
  }

  df<-do.call("cbind",results)
  df<-as.data.frame(df)
  row.names(df)<-data$Geneid
  save(df,file=rObjectFile)
} else {
  load(rObjectFile)
}

# Four libraries were resequenced in the GMP5RJ batch. Drop the superseded
# 3SQMBY originals so each sample appears exactly once. 3SQMBY_65 was a failed
# library (23MB vs ~870MB); the other three were rerun so that IL36b_96h is
# internally batch-consistent.
superseded <- c("3SQMBY_20_IL36b_96h_rep_1",
                "3SQMBY_44_IL36b_96h_rep_2",
                "3SQMBY_65_IL2_96h_rep_3",
                "3SQMBY_68_IL36b_96h_rep_3")

#df <- df[, !colnames(df) %in% superseded]

cat("\ngenes:", nrow(df), " samples:", ncol(df), "\n")
```

#### 2.2 Sample metadata

```{r design_matrix, warning = FALSE, echo=F}
#   3SQMBY_10_IL18_48h_rep_1              (batch 1)
#   GMP5RJ_1_rerun_IL36b_96h_rep_1_20     (batch 2, trailing number = original index)
sample_names <- colnames(df)

batch <- ifelse(grepl("^GMP5RJ", sample_names), "batch2", "batch1")

core <- gsub("^3SQMBY_[0-9]+_", "", sample_names)
core <- gsub("^GMP5RJ_[0-9]+_rerun_", "", core)
# only the batch 2 names carry a trailing original index; stripping it from
# batch 1 names would remove the replicate number instead
core <- ifelse(batch == "batch2", gsub("_[0-9]+$", "", core), core)

replicate <- gsub(".*_rep_", "", core)
body      <- gsub("_rep_[0-9]+$", "", core)
treatment <- gsub("_.*", "", body)
# naive samples have no timepoint in the name, they are the 0h baseline
time      <- ifelse(grepl("_", body), gsub(".*_", "", body), "0h")

design_matrix <- tibble(
  sample_name = sample_names,
  treatment   = factor(treatment, levels = names(treatment_colours)),
  time        = factor(time, levels = c("0h","24h","48h","72h","96h","120h")),
  replicate   = replicate,
  batch       = batch
) %>%
  mutate(
    group = paste(treatment, time, sep = "_"),
    label = paste(treatment, time, paste0("rep", replicate), sep = "_")
  )

row.names(design_matrix) <- design_matrix$sample_name

# stop early if a filename did not parse, rather than fitting a broken model
stopifnot(!any(is.na(design_matrix$treatment)))
stopifnot(!any(is.na(design_matrix$time)))
#stopifnot(!anyDuplicated(design_matrix$label))

# use the readable labels from here on
batch <- gsub("batch","b",batch)
colnames(df) <- paste0(design_matrix$label, "_", batch)

write.csv(design_matrix, paste0(tableDir,"sample_metadata.csv"), row.names = FALSE)

datatable(design_matrix, options = list(pageLength = 10))
```

```{r group_sizes, warning = FALSE}
# sanity check: every group should have three replicates
design_matrix %>%
  count(group, batch) %>%
  arrange(group) %>%
  as.data.frame()
```

#### 2.3 Annotate with gene symbols

```{r annotation, warning = FALSE}
# Same approach as the dane_rnaseq analysis: map Ensembl -> Entrez -> symbol,
# then collapse duplicated symbols keeping the row with the larger total count.
egENSEMBL <- toTable(org.Mm.egENSEMBL)
ids <- gsub("\\..*","",row.names(df))
row.names(df) <- ids

m <- match(row.names(df), egENSEMBL$ensembl_id)
df$EntrezGene <- egENSEMBL$gene_id[m]

egSYMBOL <- toTable(org.Mm.egSYMBOL)
m <- match(df$EntrezGene, egSYMBOL$gene_id)
df$symbol <- egSYMBOL$symbol[m]

#eliminate duplicated symbols, keep the one with a larger overall count
o <- order(rowSums(df[,1:(dim(df)[2]-2)]), decreasing=TRUE)
df <- df[o,]
d <- duplicated(df$symbol)
df <- df[!d,]
df <- df[!is.na(df$symbol),]
row.names(df) <- df$symbol

counts <- df %>%
  dplyr::select(-EntrezGene, -symbol)

# keep the metadata in the same order as the count columns
#design_matrix <- design_matrix[match(colnames(counts), design_matrix$label),]

cat("genes with symbols:", nrow(counts), "\n")
```

#### 2.4 Library sizes

```{r library_sizes, fig.width = 12, fig.height = 5}
lib_df <- design_matrix %>%
  mutate(lib_size = colSums(counts))

p <- ggplot(lib_df, aes(x = label, y = lib_size/1e6, fill = treatment)) +
  geom_col() +
  scale_fill_manual(values = treatment_colours) +
  labs(title = "Assigned reads per sample, after UMI deduplication",
       x = NULL, y = "Millions of assigned reads") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6))
p

ggsave(paste0(imageDir,"library_sizes.pdf"), p, width = 12, height = 5)
```

#### 2.5 MDS plot

```{r mds_plot, fig.width = 9, fig.height = 7}
group <- factor(design_matrix$group)

y <- DGEList(counts = counts, group = group)
keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y)

cat("genes retained after filterByExpr:", nrow(y), "\n")

mds_result <- plotMDS(y, top = 1000, plot = FALSE)

sample_df <- design_matrix %>%
  mutate(PCA1 = mds_result$x,
         PCA2 = mds_result$y)

# colour by cytokine, shape by timepoint - one colour per group would give 24
# colours and no readable structure
p <- ggplot(sample_df, aes(x = PCA1, y = PCA2, color = treatment, shape = time, label = label)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 2.5, max.overlaps = 30, show.legend = F) +
  scale_color_manual(values = treatment_colours) +
  labs(title = "MDS plot",
       x = "Dimension 1",
       y = "Dimension 2") +
  theme_minimal()
p

ggsave(paste0(imageDir,"MDS_plot.pdf"), p, width = 9, height = 7)
```

```{r mds_batch, fig.width = 9, fig.height = 7}
# the same MDS coloured by batch. If batch 2 samples sit apart from their
# batch 1 group-mates, the IL36b_96h contrasts carry a batch effect that
# cannot be separated from the cytokine effect.
p <- ggplot(sample_df, aes(x = PCA1, y = PCA2, color = batch, shape = time, label = label)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(data = subset(sample_df, batch == "batch2"),
                           size = 3, max.overlaps = 30, show.legend = F) +
  scale_color_manual(values = c("batch1" = "gray70", "batch2" = "#F20D0D")) +
  labs(title = "MDS plot, coloured by sequencing batch",
       x = "Dimension 1", y = "Dimension 2") +
  theme_minimal()
p

ggsave(paste0(imageDir,"MDS_plot_batch.pdf"), p, width = 9, height = 7)
```

### 3. Differential expression

#### 3.1 edgeR analysis

```{r edger_analysis, warnings = F, include = FALSE}
# One model fitted across all samples, then each contrast pulled out of it.
# The dane_rnaseq analysis subset to two groups per comparison, which is fine
# with three groups; here there are 24, and estimating the dispersion once from
# the whole dataset is much better informed than from six samples at a time.
group <- factor(design_matrix$group)
design <- model.matrix(~0+group)
colnames(design) <- levels(group)

y <- DGEList(counts = counts, group = group)
keep <- filterByExpr(y, design)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

cytokines <- c("IL2","IL715","IL18","IL36b")

# Contrasts are generated rather than typed out so they cannot drift out of
# step with the data. NC vs naive shows the culture effect on its own.
combinations <- list()

for(tp in c("24h","48h","72h")){
  combinations[[paste0("NC_",tp,"_vs_Naive_0h")]] <- c(paste0("NC_",tp), "Naive_0h")
}

for(cyt in cytokines){
  # time-matched against NC, the interpretable comparisons
  for(tp in c("24h","48h","72h")){
    combinations[[paste0(cyt,"_",tp,"_vs_NC_",tp)]] <- c(paste0(cyt,"_",tp), paste0("NC_",tp))
  }
  # no NC exists at these timepoints, so naive is the only available reference
  for(tp in c("96h","120h")){
    combinations[[paste0(cyt,"_",tp,"_vs_Naive_0h")]] <- c(paste0(cyt,"_",tp), "Naive_0h")
  }
}

# drop any contrast whose groups are not both present
combinations <- combinations[sapply(combinations, function(x) all(x %in% colnames(design)))]

top_tables <- list()
for(combination in names(combinations)){

  treatment1 <- combinations[[combination]][1]
  treatment2 <- combinations[[combination]][2]

  cat(combination)
  cat("\n")

  myargs = list(paste0(treatment1,"-",treatment2), levels=design)
  contrasts <- do.call(makeContrasts, myargs)
  qlf <- glmQLFTest(fit, contrast = contrasts)
  top.table <- topTags(qlf, n = Inf)
  top.table <- as.data.frame(top.table) %>%
    tibble::rownames_to_column("gene")
  write.csv(top.table,paste0(tableDir,combination,".csv"),quote=F,row.names=F)

  expression<-as.data.frame(top.table)
  row.names(expression) <- expression$gene
  top_tables[[combination]] <- expression
}
```

#### 3.2 Number of DE genes per comparison

```{r de_summary, fig.width = 12, fig.height = 6}
de_summary <- imap_dfr(top_tables, function(tbl, comp_name){
  tbl %>%
    mutate(direction = case_when(
      FDR < 0.05 & logFC >= 1  ~ "up",
      FDR < 0.05 & logFC <= -1 ~ "down",
      TRUE ~ "ns")) %>%
    filter(direction != "ns") %>%
    count(direction, name = "n_genes") %>%
    mutate(comparison = comp_name)
})

de_summary <- de_summary %>%
  mutate(comparison = factor(comparison, levels = names(combinations))) %>%
  complete(comparison, direction = c("down","up"), fill = list(n_genes = 0)) %>%
  mutate(signed_n = ifelse(direction == "down", -n_genes, n_genes))

p <- ggplot(de_summary, aes(x = comparison, y = signed_n, fill = direction)) +
  geom_col(width = 0.75, colour = "black") +
  geom_hline(yintercept = 0, colour = "black") +
  scale_fill_manual(values = c("down" = "#0096FF", "up" = "#F20D0D")) +
  labs(title = "Differentially expressed genes by comparison",
       subtitle = "FDR < 0.05, |log2FC| >= 1",
       x = "Comparison", y = "Number of DE genes") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p

ggsave(paste0(imageDir,"DE_gene_counts.pdf"), p, width = 12, height = 6)
```

#### 3.3 Volcano plots

```{r volcanoes, fig.width = 8, fig.height = 5, results = "asis"}
# T-cell differentiation markers, same list used in the dane_rnaseq volcanoes
marker_genes <- c("Batf","Nr4a1","Id3","Id2","Gzmb","Tnf","Ifng","Tcf7","Sell","Ccr7",
                  "Slamf6","Bcl2","Bach2","Tbx21","Prdm1","Il2ra","Il2rb",
                  "Cxcr3","Cxcr6","Pdcd1","Ctla4","Foxo1","Myb","Klrg1")

# looped because the contrast list is generated, so its length can change
for(comparison in names(top_tables)){
  cat("\n\n##### ", comparison, "\n\n", sep="")
  p <- plot_volcano(top_tables[[comparison]],
                    max.overlaps = 10000,
                    highlight_genes = marker_genes) +
    ggtitle(comparison)
  print(p)
  cat("\n\n")
}
```

#### 3.4 Combined DE table

```{r total_DE_tables, echo=FALSE}
table_df <- imap_dfr(top_tables, function(tbl, comp_name){
  tbl %>%
    filter(FDR < 0.05) %>%
    dplyr::select(gene, logFC, FDR) %>%
    mutate(treatment = comp_name)
})

table_df_pivot <- table_df %>%
  pivot_wider(
    id_cols = gene,
    names_from = treatment,
    values_from = c(logFC, FDR),
    names_glue = "{treatment}_{.value}",
    values_fn = dplyr::first
  ) %>%
  rename_with(~ gsub("logFC", "log2FC", .), everything()) %>%
  mutate(across(contains("log2FC"), ~ round(.x, 3))) %>%
  mutate(across(contains("FDR"), ~ signif(.x, 3)))

table_df_pivot <- table_df_pivot[, c("gene", sort(colnames(table_df_pivot)[-1]))]

datatable(table_df_pivot,
  extensions = 'Buttons',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ))
```

#### 3.5 Plotting individual genes

```{r plot_counts, fig.width = 10, fig.height = 8, warning=FALSE}
#Select genes to plot
genes_to_plot <- c("Nr4a1", "Id3", "Gzmb", "Tnf", "Ifng", "Tcf7")

dge <- DGEList(counts = counts, group = design_matrix$group)
dge <- calcNormFactors(dge, method = "TMM")
count_matrix <- cpm(dge, log = FALSE)

# long format, one row per gene x sample, with treatment and time attached
dt_long <- as.data.frame(count_matrix) %>%
  tibble::rownames_to_column("Genes") %>%
  filter(Genes %in% genes_to_plot) %>%
  pivot_longer(-Genes, names_to = "label", values_to = "Expression") %>%
  left_join(design_matrix, by = "label")

# one panel per gene, x axis is the timecourse, one line per cytokine
p <- ggplot(dt_long, aes(x = time, y = Expression, fill = treatment)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~Genes, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = treatment_colours) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  labs(y = "Gene Expression (CPM)", x = NULL) +
  theme_bw() +
  ylim(c(0,NA))
p

ggsave(paste0(imageDir,"marker_genes.pdf"), p, width = 10, height = 8)
```

### 4. Pathway analysis

#### 4.1 Hallmark enrichment

```{r pathway_analysis, fig.width = 10, fig.height = 12, warning=FALSE}
#get the hallmarks database
h_gene_sets = msigdbr(species = "mouse", category = "H")
local_gene_sets_df <- h_gene_sets %>%
  dplyr::select(gs_name, gene_symbol)

# run enrichment separately on the up and down gene lists of each contrast,
# returning NULL when a list is too short to test
run_enrichment <- function(genes, direction_label, treatment_label, min_size = 5){
  genes <- unique(na.omit(genes))
  if(length(genes) < min_size) return(NULL)

  enr <- clusterProfiler::enricher(gene = genes,
                                   TERM2GENE = local_gene_sets_df,
                                   pAdjustMethod = "BH")
  if(is.null(enr)) return(NULL)
  res <- as.data.frame(enr)
  if(nrow(res) == 0) return(NULL)

  res %>%
    mutate(direction = direction_label, treatment = treatment_label)
}

results <- list()
for(combination in names(top_tables)){
  tbl <- top_tables[[combination]]
  up   <- tbl %>% filter(FDR < 0.05, logFC >= 1)  %>% pull(gene)
  down <- tbl %>% filter(FDR < 0.05, logFC <= -1) %>% pull(gene)

  results[[paste0(combination,"_up")]]   <- run_enrichment(up,   "Upregulated",   combination)
  results[[paste0(combination,"_down")]] <- run_enrichment(down, "Downregulated", combination)
}

# Combine all enrichment results into one dataframe
collated <- bind_rows(results) %>%
  mutate(pathway = gsub("HALLMARK_", "", ID)) %>%
  mutate(
    treatment = forcats::fct_relevel(treatment, names(combinations)),
    pathway = fct_rev(fct_relevel(pathway, sort(unique(pathway))))
  ) %>%
  filter(p.adjust < 0.05)

write.csv(collated, paste0(tableDir,"hallmark_enrichment.csv"), row.names = FALSE)

p <- collated %>%
  ggplot(aes(treatment, pathway, size = FoldEnrichment)) +
  geom_point(aes(color = direction)) +
  facet_wrap(~direction, nrow = 1) +
  scale_color_manual(values = c(
    "Upregulated" = "firebrick",
    "Downregulated" = "steelblue"
  )) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(size = 8, face = "bold"),
    legend.position = "right"
  ) +
  labs(
    title = "Pathway Enrichment across Treatments",
    x = "Treatment",
    y = "Pathway"
  )
p

ggsave(paste0(imageDir,"hallmark_enrichment.pdf"), p, width = 10, height = 12)
```

#### 4.2 Enrichment table

```{r pathway_table, echo=FALSE}
datatable(
  collated %>%
    dplyr::select(treatment, direction, pathway, GeneRatio, BgRatio, p.adjust, Count) %>%
    mutate(p.adjust = signif(p.adjust, 3)) %>%
    arrange(p.adjust),
  extensions = 'Buttons',
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ))
```

### 5. Session info

```{r session_info}
sessionInfo()
```
