---
title: "Cytokine-driven T-cell differentiation timecourse"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{ben_legg_cytokine_timecourse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteBuild{true}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

### 1. Workplace setup

```{r setup, warning = FALSE, include = FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(purrr)
library(forcats)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(edgeR)
library(org.Mm.eg.db)
library(msigdbr)
library(clusterProfiler)
library(pheatmap)
library(DT)
library(patchwork)
```

#### 1.1 Directory structure

```{r directory_setup, warning = FALSE, echo=F}
# 1. make sure the bioinf_scratch drive is mounted
# 2. session -> set working directory -> scripts

######## directory structure #######
projectDir=paste0("../")
resultsDir=paste0(projectDir,"/results")
imageDir=paste0(resultsDir,"/figures/")
robjectsDir = paste(resultsDir,"/Robjects/",sep="")
tableDir = paste(resultsDir,"/tables/",sep="")

system(paste("mkdir -p",robjectsDir))
system(paste("mkdir -p",tableDir))
system(paste("mkdir -p",imageDir))

# counts come from 5.feature_counts.sbatch, one file per sample
inDir<-paste0(resultsDir,"/counts/featurecounts/")
```

#### 1.2 Experimental design

Naive T-cells cultured with one of four cytokines across a timecourse, against
a negative control (NC) and an uncultured naive baseline.

1. **NC only exists at 24h, 48h and 72h.** The 96h and 120h cytokine samples
   have no time-matched control. 
2. **Batch is handled by sample selection, not by modelling.** Four libraries
   were resequenced in a second batch (`GMP5RJ`). The three IL36b_96h reruns
   overlapped their batch 1 originals exactly, so the reruns were dropped and
   IL36b_96h is entirely batch 1. The IL2_96h rep 3 rerun is kept, because its
   batch 1 original failed and the rerun behaves like its group-mates. Batch is
   therefore not a term in the DE model. One further sample, IL715_24h rep 3,
   was removed as an outlier, so that group has two replicates.



```{r colours, warning = FALSE, echo=F}
# IL2 / IL7_15 / IL36 keep the same colours used in the dane_rnaseq analysis so
# the two Ben projects can be read side by side. IL18 is new here and takes
# orange. Naive and NC are greys, because they are references rather than
# treatments of interest.
treatment_colours <- c(
  "Naive" = "gray70",
  "NC"    = "gray40",
  "IL2"   = "#F20D0D",
  "IL715" = "#00B050",
  "IL18"  = "#FF8C00",
  "IL36b" = "#0096FF"
)

# Direction colours, shared by the volcanoes and the pathway dotplot.
direction_colours <- c(
  "DOWN" = "#0096FF",
  "NO"   = "gray60",
  "UP"   = "#F20D0D"
)
```

```{r volcano_function, include = FALSE, echo=F}
# Same volcano used in 8.ben_volcanoes.Rmd. Pass highlight_genes to label a
# fixed marker list instead of every significant gene.
plot_volcano <- function(top,
                         FDR_cutoff = 0.05,
                         max.overlaps = 30,
                         highlight_genes = NULL) {
  top <- as.data.frame(top)

  # Rename column if needed
  if ("adj.P.Val" %in% colnames(top)) {
    top <- top %>%
      rename(FDR = adj.P.Val)
  }

  # Differential expression flags. Built with ifelse rather than by indexing
  # into a subset, so a contrast with no significant genes does not error.
  top$diff_expression <- "NO"
  top$diff_expression[top$logFC >= 1  & top$FDR <= FDR_cutoff] <- "UP"
  top$diff_expression[top$logFC <= -1 & top$FDR <= FDR_cutoff] <- "DOWN"

  # Label logic
  top$labelgenes <- ""
  if (!is.null(highlight_genes)) {
    highlight_genes_present <- intersect(highlight_genes, rownames(top))
    top[rownames(top) %in% highlight_genes_present, "labelgenes"] <-
      rownames(top)[rownames(top) %in% highlight_genes_present]
  } else {
    top[top$diff_expression != "NO", "labelgenes"] <-
      rownames(top[top$diff_expression != "NO", ])
  }

  ggplot(top, aes(x = logFC, y = -log10(FDR), color = diff_expression, label = labelgenes)) +
    geom_point(size = 1.8) +
    geom_text_repel(
      max.overlaps = max.overlaps,
      min.segment.length = 0,       # always draw line
      segment.color = "grey40",     # color of connecting line
      segment.size = 0.6,           # thickness of the line
      box.padding = 0.3,
      point.padding = 0.25,
      seed = 123                    # reproducible layout
    ) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "darkred") +
    geom_hline(yintercept = -log10(FDR_cutoff), linetype = "dashed", color = "darkred") +
    scale_color_manual(values = direction_colours) +
    theme_classic(base_size = 12) +
    labs(
      x = "log2 Fold Change",
      y = "-log10(FDR)",
      color = "Differential expression",
      title = "Volcano plot"
    )
}
```

### 2. QC

#### 2.1 Import data

```{r data_import, warning = FALSE, echo=F}
rObjectFile<-paste0(robjectsDir,"featurecounts_files.Rdata")

# featureCounts writes a "# Program:.." banner, then a wide table whose LAST
# column holds the counts and is named after the input BAM path.
if(!file.exists(rObjectFile)){
  inFiles<-list.files(inDir,full.names=T, pattern=".featureCounts.txt$")
  results<-list()
  for(inFile in inFiles){

    sampleName<-basename(inFile)
    sampleName<-gsub(".featureCounts.txt","",sampleName)
    sampleName<-gsub("_trimmed","",sampleName)

    data<-read.delim(inFile,comment.char="#")

    cat(sampleName)
    cat("\n")
    results[[sampleName]]<-data[,ncol(data)]
  }

  df<-do.call("cbind",results)
  df<-as.data.frame(df)
  row.names(df)<-data$Geneid
  save(df,file=rObjectFile)
} else {
  load(rObjectFile)
}

# ---- samples excluded from every analysis ----
#
# 1. 3SQMBY_65 was a failed library (23MB against ~870MB for everything else).
#    Its batch 2 rerun GMP5RJ_3 replaces it, and that rerun sits where it should
#    on the PCA, so it is kept.
# 2. The three IL36b_96h batch 2 reruns overlap their batch 1 originals exactly
#    on the PCA, so they add nothing. Dropping them and keeping the originals
#    makes IL36b_96h entirely batch 1, which removes the batch confound in that
#    group rather than having to model around it.
# 3. 3SQMBY_52 (IL715_24h rep 3) is a PCA outlier.
#
# After these exclusions the only batch 2 sample left is IL2_96h rep 3, which
# behaves like its group-mates - which is why batch is deliberately NOT a term
# in the DE model below.
excluded_samples <- c(
  "3SQMBY_65_IL2_96h_rep_3",            # failed library, superseded by GMP5RJ_3
  "GMP5RJ_1_rerun_IL36b_96h_rep_1_20",  # duplicates 3SQMBY_20
  "GMP5RJ_2_rerun_IL36b_96h_rep_2_44",  # duplicates 3SQMBY_44
  "GMP5RJ_4_rerun_IL36b_96h_rep_3_68",  # duplicates 3SQMBY_68
  "3SQMBY_52_IL715_24h_rep_3"           # outlier
)

missing <- setdiff(excluded_samples, colnames(df))
if(length(missing) > 0){
  stop("excluded sample not found in the count matrix: ", paste(missing, collapse=", "))
}

df <- df[, !colnames(df) %in% excluded_samples]

cat("\ngenes:", nrow(df), " samples:", ncol(df), "\n")
```

#### 2.2 Sample metadata

```{r design_matrix, warning = FALSE, echo=F}
#   3SQMBY_10_IL18_48h_rep_1              (batch 1)
#   GMP5RJ_1_rerun_IL36b_96h_rep_1_20     (batch 2, trailing number = original index)
sample_names <- colnames(df)

batch <- ifelse(grepl("^GMP5RJ", sample_names), "batch2", "batch1")

core <- gsub("^3SQMBY_[0-9]+_", "", sample_names)
core <- gsub("^GMP5RJ_[0-9]+_rerun_", "", core)
# only the batch 2 names carry a trailing original index; stripping it from
# batch 1 names would remove the replicate number instead
core <- ifelse(batch == "batch2", gsub("_[0-9]+$", "", core), core)

replicate <- gsub(".*_rep_", "", core)
body      <- gsub("_rep_[0-9]+$", "", core)
treatment <- gsub("_.*", "", body)
# naive samples have no timepoint in the name, they are the 0h baseline
time      <- ifelse(grepl("_", body), gsub(".*_", "", body), "0h")

design_matrix <- tibble(
  sample_name = sample_names,
  treatment   = factor(treatment, levels = names(treatment_colours)),
  time        = factor(time, levels = c("0h","24h","48h","72h","96h","120h")),
  replicate   = replicate,
  batch       = batch
) %>%
  mutate(
    group = paste(treatment, time, sep = "_"),
    label = paste(treatment, time, paste0("rep", replicate), sep = "_")
  )

row.names(design_matrix) <- design_matrix$sample_name

# stop early if a filename did not parse, rather than fitting a broken model
stopifnot(!any(is.na(design_matrix$treatment)))
stopifnot(!any(is.na(design_matrix$time)))
stopifnot(!anyDuplicated(design_matrix$label))

# use the readable labels from here on
batch <- gsub("batch","b",batch)
colnames(df) <- paste0(design_matrix$label, "_", batch)

write.csv(design_matrix, paste0(tableDir,"sample_metadata.csv"), row.names = FALSE)

datatable(design_matrix, options = list(pageLength = 10))
```

```{r group_sizes, warning = FALSE}
# sanity check on group sizes. Every group has three replicates except
# IL715_24h, which has two after the outlier was dropped.
design_matrix %>%
  count(group, batch) %>%
  arrange(group) %>%
  as.data.frame()
```

#### 2.2b Design completeness

```{r design_heatmap, fig.width = 7, fig.height = 4}
# How many replicates exist for every treatment x timepoint combination.
# White = nothing sequenced, dark red = the most replicates present. This makes
# the holes in the design visible at a glance: naive exists only at 0h, and NC
# was only collected at 24/48/72h, which is why the 96h and 120h cytokine
# samples have no time-matched control.
design_counts <- design_matrix %>%
  count(treatment, time, name = "n") %>%
  tidyr::complete(treatment, time, fill = list(n = 0)) %>%
  # dark tiles need light text, pale tiles need dark text
  mutate(label_colour = ifelse(n > max(n)/2, "white", "black"))

p <- ggplot(design_counts, aes(x = time, y = fct_rev(treatment), fill = n)) +
  geom_tile(colour = "grey70") +
  geom_text(aes(label = n, colour = label_colour), size = 4, show.legend = FALSE) +
  scale_colour_identity() +
  scale_fill_gradient(low = "white", high = "darkred", limits = c(0, NA)) +
  labs(title = "Replicates per treatment and timepoint",
       x = "Timepoint", y = NULL, fill = "n replicates") +
  coord_fixed() +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
p

ggsave(paste0(imageDir,"design_completeness.pdf"), p, width = 7, height = 4)
```

#### 2.3 Annotate with gene symbols

```{r annotation, warning = FALSE}
# Same approach as the dane_rnaseq analysis: map Ensembl -> Entrez -> symbol,
# then collapse duplicated symbols keeping the row with the larger total count.
egENSEMBL <- toTable(org.Mm.egENSEMBL)
ids <- gsub("\\..*","",row.names(df))
row.names(df) <- ids

m <- match(row.names(df), egENSEMBL$ensembl_id)
df$EntrezGene <- egENSEMBL$gene_id[m]

egSYMBOL <- toTable(org.Mm.egSYMBOL)
m <- match(df$EntrezGene, egSYMBOL$gene_id)
df$symbol <- egSYMBOL$symbol[m]

#eliminate duplicated symbols, keep the one with a larger overall count
o <- order(rowSums(df[,1:(dim(df)[2]-2)]), decreasing=TRUE)
df <- df[o,]
d <- duplicated(df$symbol)
df <- df[!d,]
df <- df[!is.na(df$symbol),]
row.names(df) <- df$symbol

counts <- df %>%
  dplyr::select(-EntrezGene, -symbol)

# keep the metadata in the same order as the count columns
#design_matrix <- design_matrix[match(colnames(counts), design_matrix$label),]

cat("genes with symbols:", nrow(counts), "\n")
```

#### 2.4 Library sizes

```{r library_sizes, fig.width = 12, fig.height = 5}
lib_df <- design_matrix %>%
  mutate(lib_size = colSums(counts))

p <- ggplot(lib_df, aes(x = label, y = lib_size/1e6, fill = treatment)) +
  geom_col() +
  scale_fill_manual(values = treatment_colours) +
  labs(title = "Assigned reads per sample, after UMI deduplication",
       x = NULL, y = "Millions of assigned reads") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6))
p

ggsave(paste0(imageDir,"library_sizes.pdf"), p, width = 12, height = 5)
```

#### 2.5 MDS plot

```{r mds_plot, fig.width = 9, fig.height = 7}
group <- factor(design_matrix$group)

y <- DGEList(counts = counts, group = group)
keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y)

cat("genes retained after filterByExpr:", nrow(y), "\n")

mds_result <- plotMDS(y, top = 1000, plot = FALSE)

sample_df <- design_matrix %>%
  mutate(PCA1 = mds_result$x,
         PCA2 = mds_result$y)

# colour by cytokine, shape by timepoint - one colour per group would give 24
# colours and no readable structure
p <- ggplot(sample_df, aes(x = PCA1, y = PCA2, color = treatment, shape = time, label = label)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 2.5, max.overlaps = 30, show.legend = F) +
  scale_color_manual(values = treatment_colours) +
  labs(title = "MDS plot",
       x = "Dimension 1",
       y = "Dimension 2") +
  theme_minimal()
p

ggsave(paste0(imageDir,"MDS_plot.pdf"), p, width = 9, height = 7)
```

```{r mds_batch, fig.width = 9, fig.height = 7}
# the same MDS coloured by batch. If batch 2 samples sit apart from their
# batch 1 group-mates, the IL36b_96h contrasts carry a batch effect that
# cannot be separated from the cytokine effect.
p <- ggplot(sample_df, aes(x = PCA1, y = PCA2, color = batch, shape = time, label = label)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(data = subset(sample_df, batch == "batch2"),
                           size = 3, max.overlaps = 30, show.legend = F) +
  scale_color_manual(values = c("batch1" = "gray70", "batch2" = "#F20D0D")) +
  labs(title = "MDS plot, coloured by sequencing batch",
       x = "Dimension 1", y = "Dimension 2") +
  theme_minimal()
p

ggsave(paste0(imageDir,"MDS_plot_batch.pdf"), p, width = 9, height = 7)
```

#### 2.6 PCA trajectories

```{r pca_function, warning = FALSE}
n_var_genes <- 2000

# Eta squared: the share of a component's variance explained by one factor.
# PC1 is usually dominated by time, but the axis separating the cytokines is
# often further down, so both axes are chosen by this rather than assumed.
eta_sq <- function(values, grouping){
  grouping <- factor(grouping)
  grand  <- mean(values)
  ss_tot <- sum((values - grand)^2)
  ss_bet <- sum(tapply(values, grouping, function(x) length(x) * (mean(x) - grand)^2))
  if(ss_tot == 0) return(0)
  ss_bet / ss_tot
}

# PCA is recomputed on whatever subset is passed in, rather than the full PCA
# being filtered down. A PCA fitted to one timepoint finds the axes that
# separate the cytokines at that timepoint, which is the point of subsetting.
make_pca <- function(keep, title, file, arrows = TRUE, colour_gene = NULL,
                     label_time = FALSE, method = c("pca","mds")){
  method <- match.arg(method)
  dm  <- design_matrix[keep, , drop = FALSE]
  cnt <- counts[, keep, drop = FALSE]
  if(nrow(dm) < 3) return(NULL)

  d <- DGEList(counts = cnt, group = dm$group)
  k <- filterByExpr(d, group = dm$group)
  d <- d[k, , keep.lib.sizes = FALSE]
  d <- calcNormFactors(d)
  lc <- log2(cpm(d, log = FALSE) + 1)

  gene_vars <- apply(lc, 1, var)
  top_var <- names(sort(gene_vars, decreasing = TRUE))[1:min(n_var_genes, length(gene_vars))]

  if(method == "pca"){
    # PCA on the most variable genes; components carry a % of total variance
    ord    <- prcomp(t(lc[top_var, ]), center = TRUE, scale. = FALSE)
    coords <- ord$x
    pv     <- 100 * ord$sdev^2 / sum(ord$sdev^2)
    ax_name <- "PC"
  } else {
    # edgeR/limma MDS: distances are the leading log fold changes between each
    # pair of samples, computed on the top genes for that pair. With few samples
    # this is usually better behaved than PCA, which is driven by whichever
    # genes happen to be most variable overall.
    #
    # The coordinates have to be reconstructed: limma 3.64 dropped the
    # cmdscale.out component the MDS object used to carry. Its plotted x and y
    # are eigen.vectors scaled by sqrt(eigen.values), checked exact against
    # md$x and md$y. Negative eigenvalues can occur, so only positive ones are
    # used for the dimensions offered to the axis selection below.
    md  <- plotMDS(d, top = 500, plot = FALSE)
    eva <- md$eigen.values
    ev  <- md$eigen.vectors
    k   <- max(2, min(8, sum(eva > 0)))
    coords <- sweep(ev[, seq_len(k), drop = FALSE], 2, sqrt(eva[seq_len(k)]), "*")
    pv      <- 100 * md$var.explained[seq_len(k)]
    ax_name <- "Leading logFC dim "
  }

  npc    <- min(10, ncol(coords))
  n_time <- length(unique(as.character(dm$time)))
  draw_arrows <- arrows && n_time > 1

  eta_time <- sapply(seq_len(npc), function(i) eta_sq(coords[,i], dm$time))
  eta_trt  <- sapply(seq_len(npc), function(i) eta_sq(coords[,i], dm$treatment))

  # axis label, omitting the variance term when the method does not define one
  mk_lab <- function(i, eta_val, what){
    if(is.na(pv[i])) sprintf("%s%d (%s eta2=%.2f)", ax_name, i, what, eta_val)
    else sprintf("%s%d (%.1f%% variance; %s eta2=%.2f)", ax_name, i, pv[i], what, eta_val)
  }

  if(draw_arrows){
    ax <- which.max(eta_time)
    ay <- which.max(replace(eta_trt, ax, -Inf))
    xlab <- mk_lab(ax, eta_time[ax], "time axis,")
    ylab <- mk_lab(ay, eta_trt[ay],  "condition axis,")
  } else {
    # a single timepoint has no time axis to find, so use the first two dimensions
    ax <- 1; ay <- 2
    xlab <- mk_lab(1, eta_trt[1], "condition")
    ylab <- mk_lab(2, eta_trt[2], "condition")
  }

  df <- dm %>% mutate(x = coords[, ax], y = coords[, ay])

  # numeric hours, used as the on-point label. "0" is naive.
  df$time_num <- as.numeric(gsub("h", "", as.character(df$time)))

  # colour_gene overlays one gene's expression on the same coordinates. Matched
  # case-insensitively because mouse symbols are Batf, not BATF.
  if(!is.null(colour_gene)){
    gi <- match(tolower(colour_gene), tolower(rownames(lc)))
    if(is.na(gi)) stop("gene not found in the filtered expression matrix: ", colour_gene)
    df$gene_expr <- as.numeric(lc[gi, ])
  }

  mn <- df %>%
    group_by(treatment, time) %>%
    summarise(x = mean(x), y = mean(y), .groups = "drop") %>%
    arrange(treatment, time)

  # convex hull round the replicates of each treatment x timepoint.
  # stat_ellipse is unreliable at n = 3, a hull is not.
  hl <- df %>%
    group_by(treatment, time) %>%
    filter(n() >= 3) %>%
    slice(chull(x, y)) %>%
    ungroup()

  if(is.null(colour_gene)){
    # standard version: one colour per treatment
    g <- ggplot() +
      geom_polygon(data = hl, aes(x, y, fill = treatment, group = interaction(treatment, time)),
                   alpha = 0.15, colour = NA) +
      geom_point(data = df, aes(x, y, colour = treatment), size = 2.2, alpha = 0.9)

    if(draw_arrows){
      g <- g + geom_path(data = mn, aes(x, y, colour = treatment, group = treatment),
                         linewidth = 0.7,
                         arrow = grid::arrow(length = grid::unit(0.18, "cm"), type = "closed"))
    }

    g <- g +
      scale_colour_manual(values = treatment_colours) +
      scale_fill_manual(values = treatment_colours)

  } else {
    # gene version: treatment identity drops back to grey so the only thing
    # carrying colour is the expression of that one gene
    g <- ggplot() +
      geom_polygon(data = hl, aes(x, y, group = interaction(treatment, time)),
                   fill = "grey75", alpha = 0.25, colour = NA) +
      geom_point(data = df, aes(x, y, colour = gene_expr), size = 2.8)

    if(draw_arrows){
      g <- g + geom_path(data = mn, aes(x, y, group = treatment),
                         colour = "grey55", linewidth = 0.5,
                         arrow = grid::arrow(length = grid::unit(0.18, "cm"), type = "closed"))
    }

    g <- g +
      scale_colour_gradient(low = "grey88", high = "darkred",
                            name = paste0(colour_gene, "\nlog2(CPM+1)"))
  }

  # small number above each dot saying which timepoint it is, so the order of
  # the trajectory can be read off the points themselves rather than inferred
  # from the arrow directions
  if(label_time){
    g <- g + geom_text(data = df, aes(x, y, label = time_num),
                       size = 2.2, vjust = -0.9, colour = "grey15",
                       show.legend = FALSE, inherit.aes = FALSE)
  }

  g <- g +
    labs(title = title,
         subtitle = if(draw_arrows)
           "Shaded regions are replicates at each timepoint; arrows connect condition means"
         else
           "Shaded regions are replicates; single timepoint, so no trajectory",
         x = xlab, y = ylab, colour = "Treatment", fill = "Treatment") +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  ggsave(paste0(imageDir, file), g, width = 12, height = 7)

  eta_df <- data.frame(PC = paste0(ifelse(method == "pca", "PC", "dim"), seq_len(npc)),
                       pct_var = round(pv[seq_len(npc)], 1),
                       eta_time = round(eta_time, 2),
                       eta_treatment = round(eta_trt, 2))

  list(plot = g, eta = eta_df, n = nrow(dm))
}
```

##### 2.6.1 All samples

```{r pca_all, fig.width = 12, fig.height = 7}
res <- make_pca(rep(TRUE, nrow(design_matrix)),
                "Sample movement over time (PCA)", "PCA_trajectory.pdf")
res$eta
res$plot
```

##### 2.6.2 Without naive

```{r pca_no_naive, fig.width = 12, fig.height = 7}
# naive sits far from everything else and is a single timepoint with no
# trajectory, so dropping it lets the cultured conditions spread out
res <- make_pca(design_matrix$treatment != "Naive",
                "PCA, naive excluded", "PCA_trajectory_no_naive.pdf")
res$plot
```

##### 2.6.3 Without NC

```{r pca_no_nc, fig.width = 12, fig.height = 7}
# removes the untreated culture control, so the axes are driven by differences
# between the cytokines rather than by treated versus untreated
res <- make_pca(design_matrix$treatment != "NC",
                "PCA, NC excluded", "PCA_trajectory_no_NC.pdf")
res$plot
```

##### 2.6.4 24h only

```{r pca_24h, fig.width = 12, fig.height = 7}
# one timepoint, so no trajectory: this asks purely how far apart the
# cytokines are from each other at 24h
res <- make_pca(design_matrix$time == "24h",
                "PCA, 24h samples only", "PCA_24h.pdf", arrows = FALSE)
res$eta
res$plot
```

##### 2.6.5 120h only

```{r pca_120h, fig.width = 12, fig.height = 7}
# the same question at the end of the timecourse. NC does not exist at 120h,
# so this is cytokines only.
res <- make_pca(design_matrix$time == "120h",
                "PCA, 120h samples only", "PCA_120h.pdf", arrows = FALSE)
res$eta
res$plot
```

##### 2.6.6 Sample movement with timepoints labelled

```{r pca_labelled, fig.width = 12, fig.height = 7, results = "asis"}
# The trajectory plots again, with each dot carrying its timepoint in hours.
# Naive is 0. This makes the order along each path readable directly, instead
# of having to follow the arrowheads.
labelled_subsets <- list(
  "all samples"   = list(keep = rep(TRUE, nrow(design_matrix)),     file = "PCA_trajectory_labelled.pdf"),
  "naive removed" = list(keep = design_matrix$treatment != "Naive", file = "PCA_trajectory_no_naive_labelled.pdf"),
  "NC removed"    = list(keep = design_matrix$treatment != "NC",    file = "PCA_trajectory_no_NC_labelled.pdf")
)

for(nm in names(labelled_subsets)){
  spec <- labelled_subsets[[nm]]
  cat("\n\n###### ", nm, "\n\n", sep="")
  res <- make_pca(spec$keep,
                  paste0("Sample movement over time, ", nm),
                  spec$file, arrows = TRUE, label_time = TRUE)
  if(is.null(res)){ cat("Not enough samples.\n\n"); next }
  print(res$plot)
  cat("\n\n")
}
```

##### 2.6.7 The same PCAs coloured by Batf expression

```{r pca_batf, fig.width = 12, fig.height = 7, results = "asis"}
# A PCA cannot be computed on a single gene - one gene is one dimension - so
# these are the same five decompositions with Batf's normalised expression
# overlaid on the points instead of treatment identity.
colour_gene <- "Batf"

pca_subsets <- list(
  "all samples"   = list(keep = rep(TRUE, nrow(design_matrix)),      arrows = TRUE,  file = "PCA_Batf_all.pdf"),
  "naive removed" = list(keep = design_matrix$treatment != "Naive",  arrows = TRUE,  file = "PCA_Batf_no_naive.pdf"),
  "NC removed"    = list(keep = design_matrix$treatment != "NC",     arrows = TRUE,  file = "PCA_Batf_no_NC.pdf"),
  "24h only"      = list(keep = design_matrix$time == "24h",         arrows = FALSE, file = "PCA_Batf_24h.pdf"),
  "120h only"     = list(keep = design_matrix$time == "120h",        arrows = FALSE, file = "PCA_Batf_120h.pdf")
)

for(nm in names(pca_subsets)){
  spec <- pca_subsets[[nm]]
  cat("\n\n###### Batf, ", nm, "\n\n", sep="")
  # label_time follows arrows: the multi-timepoint panels get the hours printed
  # on each dot, the single-timepoint ones do not, since every point there would
  # carry the same number
  res <- make_pca(spec$keep,
                  paste0("Batf expression on PCA, ", nm),
                  spec$file, arrows = spec$arrows, colour_gene = colour_gene,
                  label_time = spec$arrows)
  if(is.null(res)){ cat("Not enough samples.\n\n"); next }
  print(res$plot)
  cat("\n\n")
}
```

##### 2.6.8 24h and 120h, cleaner views

```{r pca_mds_alternatives, fig.width = 12, fig.height = 7, results = "asis"}
# The single-timepoint PCAs are hard to read: with 11-14 samples PCA is driven
# by whichever genes happen to be most variable, and NC sits far from the
# cytokines and stretches the axes. Two remedies, shown side by side:
#   - drop NC, so the axes are set by differences between the cytokines
#   - use edgeR's MDS instead, whose distances are leading log fold changes
#     between sample pairs and which behaves better at small n
#
# Note there is no "120h without NC" panel: NC stops at 72h, so the existing
# 120h plot already contains no NC.
alt_views <- list(
  list(nm = "24h, NC removed (PCA)",
       keep = design_matrix$time == "24h" & design_matrix$treatment != "NC",
       method = "pca", file = "PCA_24h_no_NC.pdf"),
  list(nm = "24h, all samples (MDS)",
       keep = design_matrix$time == "24h",
       method = "mds", file = "MDS_24h.pdf"),
  list(nm = "24h, NC removed (MDS)",
       keep = design_matrix$time == "24h" & design_matrix$treatment != "NC",
       method = "mds", file = "MDS_24h_no_NC.pdf"),
  list(nm = "120h, all samples (MDS)",
       keep = design_matrix$time == "120h",
       method = "mds", file = "MDS_120h.pdf")
)

for(v in alt_views){
  cat("\n\n###### ", v$nm, "\n\n", sep="")
  res <- make_pca(v$keep, v$nm, v$file, arrows = FALSE, method = v$method)
  if(is.null(res)){ cat("Not enough samples.\n\n"); next }
  print(res$plot)
  cat("\n\n")
}
```

##### 2.6.9 120h samples, coloured by individual genes

```{r pca_120h_genes, fig.width = 12, fig.height = 7, results = "asis"}
# The 120h-only PCA again, but coloured by single genes rather than treatment.
# At the final timepoint the question is whether the cytokines have separated,
# and whether that separation tracks the stemness/memory axis - Tcf7 and Id3
# are the two markers usually used to read that.
#
# Coordinates are the 120h PCA, recomputed on those samples only, so the axes
# are the ones that separate the cytokines at 120h rather than axes inherited
# from the full timecourse.
genes_on_pca_120h <- c("Tcf7", "Id3")

for(g in genes_on_pca_120h){
  cat("\n\n###### ", g, ", 120h samples\n\n", sep="")

  # a gene filtered out by filterByExpr should not kill the whole document
  res <- tryCatch(
    make_pca(design_matrix$time == "120h",
             paste0(g, " expression on PCA, 120h samples only"),
             paste0("PCA_120h_", g, ".pdf"),
             arrows = FALSE, colour_gene = g),
    error = function(e){ cat("Could not plot ", g, ": ", conditionMessage(e), "\n\n", sep=""); NULL }
  )

  if(is.null(res)) next
  print(res$plot)
  cat("\n\n")
}
```

### 3. Differential expression

#### 3.1 edgeR model and the four contrast sets

```{r edger_analysis, warnings = F, include = FALSE}
# One model fitted across all samples. All four contrast sets are pulled out of
# this single fit, so they share one dispersion estimate and stay comparable.
group <- factor(design_matrix$group)
design <- model.matrix(~0+group)
colnames(design) <- levels(group)

y <- DGEList(counts = counts, group = group)
keep <- filterByExpr(y, design)
y <- y[keep, , keep.lib.sizes = FALSE]
y <- calcNormFactors(y)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

cytokines      <- c("IL2","IL715","IL18","IL36b")
timepoints_all <- c("24h","48h","72h","96h","120h")
timepoints_nc  <- c("24h","48h","72h")   # the only timepoints that have an NC

for(d in c("vs_naive","vs_NC","IL18_IL36b_vs_IL2","IL18_IL36b_vs_IL2_IL715")){
  system(paste("mkdir -p", paste0(tableDir,d)))
}

# Each contrast is stored as the makeContrasts expression plus the list of
# groups it needs. Storing the expression rather than a pair of names is what
# lets sets 3 and 4 average two groups on one side of the comparison.
mk <- function(expr, groups) list(expr = expr, groups = groups)

# runs one contrast off the shared fit, writes its table, returns it
run_contrast <- function(entry, name, outdir){
  myargs    <- list(entry$expr, levels = design)
  contrasts <- do.call(makeContrasts, myargs)
  qlf       <- glmQLFTest(fit, contrast = contrasts)
  top.table <- as.data.frame(topTags(qlf, n = Inf)) %>%
    tibble::rownames_to_column("gene")
  write.csv(top.table, paste0(outdir,"/",name,".csv"), quote = F, row.names = F)
  row.names(top.table) <- top.table$gene
  top.table
}

# runs a whole set and returns the list of toptables
run_set <- function(combinations, outdir){
  out <- list()
  for(nm in names(combinations)){
    cat(nm); cat("\n")
    out[[nm]] <- run_contrast(combinations[[nm]], nm, outdir)
  }
  out
}

# ---- Set 1: every cytokine timepoint against the naive 0h baseline ----
# NC is not used. All five timepoints share one reference so they are directly
# comparable, but the cytokine effect is confounded with time in culture,
# because naive is an uncultured 0h sample.
combinations_naive <- list()
for(cyt in cytokines){
  for(tp in timepoints_all){
    g <- paste0(cyt,"_",tp)
    combinations_naive[[paste0(cyt,"_",tp,"_vs_Naive_0h")]] <-
      mk(paste0(g,"-Naive_0h"), c(g,"Naive_0h"))
  }
}

# ---- Set 2: the culture effect, then cytokine against its time-matched NC ----
# NC vs naive isolates what culture alone does; cytokine vs NC then gives the
# cytokine effect with that culture effect controlled for, but only at the
# three timepoints where an NC exists.
combinations_nc <- list()
for(tp in timepoints_nc){
  combinations_nc[[paste0("NC_",tp,"_vs_Naive_0h")]] <-
    mk(paste0("NC_",tp,"-Naive_0h"), c(paste0("NC_",tp),"Naive_0h"))
}
for(cyt in cytokines){
  for(tp in timepoints_nc){
    g <- paste0(cyt,"_",tp); nc <- paste0("NC_",tp)
    combinations_nc[[paste0(cyt,"_",tp,"_vs_NC_",tp)]] <- mk(paste0(g,"-",nc), c(g,nc))
  }
}

# ---- Set 3: IL18 + IL36b combined, against IL2 ----
# The two sides are the average of the two IL18/IL36b group means versus the
# IL2 group mean at the same timepoint. Because both sides are cultured for the
# same length of time, this needs no naive or NC reference at all - which is
# why it is available at all five timepoints, unlike set 1.
combinations_combo_il2 <- list()
for(tp in timepoints_all){
  a <- paste0("IL18_",tp); b <- paste0("IL36b_",tp); c2 <- paste0("IL2_",tp)
  combinations_combo_il2[[paste0("IL18_IL36b_vs_IL2_",tp)]] <-
    mk(paste0("(",a,"+",b,")/2-",c2), c(a,b,c2))
}

# ---- Set 4: IL18 + IL36b combined, against IL2 + IL7/15 combined ----
# Same idea, but the reference is the average of the two IL2/IL715 group means.
combinations_combo_pair <- list()
for(tp in timepoints_all){
  a <- paste0("IL18_",tp); b <- paste0("IL36b_",tp)
  c2 <- paste0("IL2_",tp); d2 <- paste0("IL715_",tp)
  combinations_combo_pair[[paste0("IL18_IL36b_vs_IL2_IL715_",tp)]] <-
    mk(paste0("(",a,"+",b,")/2-(",c2,"+",d2,")/2"), c(a,b,c2,d2))
}

# drop any contrast whose groups are not all present, so a partial dataset still runs
keep_available <- function(x) x[sapply(x, function(e) all(e$groups %in% colnames(design)))]
combinations_naive      <- keep_available(combinations_naive)
combinations_nc         <- keep_available(combinations_nc)
combinations_combo_il2  <- keep_available(combinations_combo_il2)
combinations_combo_pair <- keep_available(combinations_combo_pair)

top_tables_naive      <- run_set(combinations_naive,      paste0(tableDir,"vs_naive"))
top_tables_nc         <- run_set(combinations_nc,         paste0(tableDir,"vs_NC"))
top_tables_combo_il2  <- run_set(combinations_combo_il2,  paste0(tableDir,"IL18_IL36b_vs_IL2"))
top_tables_combo_pair <- run_set(combinations_combo_pair, paste0(tableDir,"IL18_IL36b_vs_IL2_IL715"))

# all four sets in one place so the plotting chunks below can loop over them
top_tables_sets <- list(
  "1_vs_naive"              = top_tables_naive,
  "2_vs_NC"                 = top_tables_nc,
  "3_IL18_IL36b_vs_IL2"     = top_tables_combo_il2,
  "4_IL18_IL36b_vs_IL2IL715" = top_tables_combo_pair
)
combinations_sets <- list(
  "1_vs_naive"              = combinations_naive,
  "2_vs_NC"                 = combinations_nc,
  "3_IL18_IL36b_vs_IL2"     = combinations_combo_il2,
  "4_IL18_IL36b_vs_IL2IL715" = combinations_combo_pair
)

# short folder names for the per-set outputs
set_files <- c("1_vs_naive"               = "vs_naive",
               "2_vs_NC"                  = "vs_NC",
               "3_IL18_IL36b_vs_IL2"      = "IL18_IL36b_vs_IL2",
               "4_IL18_IL36b_vs_IL2IL715" = "IL18_IL36b_vs_IL2_IL715")

# One folder of volcanoes per set, plus a folder for the DE count barcharts.
# 45 volcano PDFs in a single directory is unusable; split by set they are
# straightforward to page through.
system(paste("mkdir -p", paste0(imageDir,"DE_counts")))
for(f in set_files){
  system(paste("mkdir -p", paste0(imageDir,"volcanoes/",f)))
}
```

```{r set_summary}
# what each set contains, and which object holds it
data.frame(
  set = c("1: each cytokine vs Naive 0h",
          "2: NC vs naive, then cytokine vs time-matched NC",
          "3: IL18+IL36b vs IL2",
          "4: IL18+IL36b vs IL2+IL715"),
  object = c("top_tables_naive","top_tables_nc",
             "top_tables_combo_il2","top_tables_combo_pair"),
  n_contrasts = sapply(top_tables_sets, length)
)
```

#### 3.2 Number of DE genes per comparison

```{r de_summary, fig.width = 15, fig.height = 8}
# counts up and down genes for every contrast across all four sets
de_summary <- imap_dfr(top_tables_sets, function(tt, set_name){
  imap_dfr(tt, function(tbl, comp_name){
    tbl %>%
      mutate(direction = case_when(
        FDR < 0.05 & logFC >= 1  ~ "up",
        FDR < 0.05 & logFC <= -1 ~ "down",
        TRUE ~ "ns")) %>%
      filter(direction != "ns") %>%
      count(direction, name = "n_genes") %>%
      mutate(comparison = comp_name, set = set_name)
  })
})

# make sure a contrast with zero DE genes in both directions still gets a slot
all_pairs <- imap_dfr(top_tables_sets, function(tt, set_name){
  expand.grid(comparison = names(tt), direction = c("down","up"),
              set = set_name, stringsAsFactors = FALSE)
})

de_summary <- all_pairs %>%
  left_join(de_summary, by = c("comparison","direction","set")) %>%
  mutate(n_genes = ifelse(is.na(n_genes), 0, n_genes),
         signed_n = ifelse(direction == "down", -n_genes, n_genes),
         comparison = factor(comparison, levels = unlist(lapply(top_tables_sets, names))))

# the same barchart is drawn once per set and once combined, so it is defined
# as a function rather than repeated
plot_de_counts <- function(dat, title){
  ggplot(dat, aes(x = comparison, y = signed_n, fill = direction)) +
    geom_col(width = 0.75, colour = "black") +
    geom_hline(yintercept = 0, colour = "black") +
    scale_fill_manual(values = c("down" = "#0096FF", "up" = "#F20D0D")) +
    labs(title = title,
         subtitle = "FDR < 0.05, |log2FC| >= 1",
         x = "Comparison", y = "Number of DE genes") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
}

# combined overview across all four sets
p_all <- plot_de_counts(de_summary, "Differentially expressed genes by comparison") +
  facet_wrap(~set, scales = "free_x", nrow = 2) +
  theme(axis.text.x = element_text(size = 7))
p_all

ggsave(paste0(imageDir,"DE_counts/DE_gene_counts_all_sets.pdf"), p_all, width = 15, height = 8)
```

##### 3.2.1 One barchart per set

```{r de_summary_per_set, fig.width = 10, fig.height = 6, results = "asis"}
# each set on its own axis, so the smaller sets are not squashed by set 1
for(set_name in names(top_tables_sets)){
  cat("\n\n###### Set ", set_name, "\n\n", sep="")
  dat <- de_summary %>%
    filter(set == set_name) %>%
    droplevels()
  p <- plot_de_counts(dat, paste("DE genes, set", set_name))
  print(p)
  ggsave(paste0(imageDir,"DE_counts/DE_gene_counts_", set_files[[set_name]], ".pdf"),
         p, width = 10, height = 6)
  cat("\n\n")
}
```

##### 3.2.2 Set 1 grouped by timepoint

```{r de_summary_set1_by_time, fig.width = 9, fig.height = 12}
# The same set 1 counts, reordered: all four cytokines at 24h, then all four at
# 48h, and so on - rather than walking one cytokine through the whole
# timecourse before starting the next. Timepoint becomes the outer grouping,
# so the question "does the response grow or shrink over the timecourse" reads
# straight across the panels.
#
# Treatment and timepoint are taken from the contrast definitions rather than
# parsed back out of the display names, so they cannot drift apart.
first_group <- sapply(combinations_naive, function(e) e$groups[1])
set1_key <- data.frame(
  comparison = names(combinations_naive),
  trt = sub("_.*$", "", first_group),
  tp  = sub("^[^_]+_", "", first_group),
  stringsAsFactors = FALSE
)

dat_s1 <- de_summary %>%
  filter(set == "1_vs_naive") %>%
  mutate(comparison = as.character(comparison)) %>%
  left_join(set1_key, by = "comparison") %>%
  mutate(trt = factor(trt, levels = cytokines),
         tp  = factor(tp,  levels = timepoints_all))

stopifnot(!any(is.na(dat_s1$trt)), !any(is.na(dat_s1$tp)))

p <- ggplot(dat_s1, aes(x = trt, y = signed_n, fill = direction)) +
  geom_col(width = 0.75, colour = "black") +
  geom_hline(yintercept = 0, colour = "black") +
  facet_grid(~ tp, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c("down" = "#0096FF", "up" = "#F20D0D")) +
  labs(title = "DE genes, set 1, grouped by timepoint",
       subtitle = "FDR < 0.05, |log2FC| >= 1. Each panel is one timepoint; bars are the four cytokines against naive 0h",
       x = NULL, y = "Number of DE genes") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        panel.spacing = grid::unit(0.6, "lines"),
        strip.background = element_rect(fill = "grey92"),
        panel.grid.major.x = element_blank())
p

ggsave(paste0(imageDir,"DE_counts/DE_gene_counts_set1_by_timepoint.pdf"),
       p, width = 9, height = 12)
```

#### 3.3 Volcano plots

```{r marker_genes_list, include = FALSE}
# T-cell differentiation markers, same list used in the dane_rnaseq volcanoes
marker_genes <- c("Batf","Nr4a1","Id3","Id2","Gzmb","Tnf","Ifng","Tcf7","Sell","Ccr7",
                  "Slamf6","Bcl2","Bach2","Tbx21","Prdm1","Il2ra","Il2rb",
                  "Cxcr3","Cxcr6","Pdcd1","Ctla4","Foxo1","Myb","Klrg1")
```

```{r volcanoes, fig.width = 8, fig.height = 5, results = "asis"}
# one heading per set, then one volcano per contrast within it
for(set_name in names(top_tables_sets)){
  cat("\n\n##### Set ", set_name, "\n\n", sep="")
  tt <- top_tables_sets[[set_name]]
  for(comparison in names(tt)){
    cat("\n\n###### ", comparison, "\n\n", sep="")
    p <- plot_volcano(tt[[comparison]],
                      max.overlaps = 10000,
                      highlight_genes = marker_genes) +
      ggtitle(comparison)
    print(p)
    # one PDF per contrast, filed under its set
    ggsave(paste0(imageDir,"volcanoes/",set_files[[set_name]],"/",comparison,".pdf"),
           p, width = 8, height = 5)
    cat("\n\n")
  }
}
```

#### 3.4 Combined DE tables

```{r de_table_function, include = FALSE}
# one wide table per set: a log2FC and an FDR column for every contrast
make_de_table <- function(top_tables){
  imap_dfr(top_tables, function(tbl, comp_name){
    tbl %>%
      filter(FDR < 0.05) %>%
      dplyr::select(gene, logFC, FDR) %>%
      mutate(treatment = comp_name)
  }) %>%
    pivot_wider(
      id_cols = gene,
      names_from = treatment,
      values_from = c(logFC, FDR),
      names_glue = "{treatment}_{.value}",
      values_fn = dplyr::first
    ) %>%
    rename_with(~ gsub("logFC", "log2FC", .), everything()) %>%
    mutate(across(contains("log2FC"), ~ round(.x, 3))) %>%
    mutate(across(contains("FDR"), ~ signif(.x, 3)))
}

show_de_table <- function(tab){
  if(nrow(tab) == 0) return(invisible(NULL))
  tab <- tab[, c("gene", sort(colnames(tab)[-1]))]
  datatable(tab, extensions = 'Buttons',
    options = list(pageLength = 10, scrollX = TRUE, dom = 'Bfrtip',
                   buttons = c('copy','csv','excel')))
}
```

##### 3.4.1 Set 1, versus naive 0h

```{r de_table_1, echo=FALSE}
de_table_naive <- make_de_table(top_tables_naive)
write.csv(de_table_naive, paste0(tableDir,"DE_table_vs_naive.csv"), row.names = FALSE)
show_de_table(de_table_naive)
```

##### 3.4.2 Set 2, versus time-matched NC

```{r de_table_2, echo=FALSE}
de_table_nc <- make_de_table(top_tables_nc)
write.csv(de_table_nc, paste0(tableDir,"DE_table_vs_NC.csv"), row.names = FALSE)
show_de_table(de_table_nc)
```

##### 3.4.3 Set 3, IL18+IL36b versus IL2

```{r de_table_3, echo=FALSE}
de_table_combo_il2 <- make_de_table(top_tables_combo_il2)
write.csv(de_table_combo_il2, paste0(tableDir,"DE_table_IL18_IL36b_vs_IL2.csv"), row.names = FALSE)
show_de_table(de_table_combo_il2)
```

##### 3.4.4 Set 4, IL18+IL36b versus IL2+IL7/15

```{r de_table_4, echo=FALSE}
de_table_combo_pair <- make_de_table(top_tables_combo_pair)
write.csv(de_table_combo_pair, paste0(tableDir,"DE_table_IL18_IL36b_vs_IL2_IL715.csv"), row.names = FALSE)
show_de_table(de_table_combo_pair)
```

#### 3.5 Plotting individual genes

```{r gene_plot_function, include = FALSE}
# Expression matrix built once and reused by every gene plot below.
# log2(CPM + 1): TMM-normalised CPM, then logged so fold changes read as equal
# vertical distances. The +1 keeps genes with zero counts finite rather than -Inf.
dge <- DGEList(counts = counts, group = design_matrix$group)
dge <- calcNormFactors(dge, method = "TMM")
count_matrix <- log2(cpm(dge, log = FALSE) + 1)

# Join on the actual count column names rather than on design_matrix$label.
# The count columns carry the _b1/_b2 batch suffix and the labels do not, so a
# join on "label" silently produces NAs.
sample_key <- design_matrix %>%
  mutate(column = colnames(counts))

# one panel per gene, x axis is the timecourse, one box per cytokine
plot_genes <- function(genes, title, ncol = 3){
  genes <- intersect(genes, rownames(count_matrix))
  if(length(genes) == 0) return(NULL)

  dt_long <- as.data.frame(count_matrix[genes, , drop = FALSE]) %>%
    tibble::rownames_to_column("Genes") %>%
    pivot_longer(-Genes, names_to = "column", values_to = "Expression") %>%
    left_join(sample_key, by = "column")

  stopifnot(!any(is.na(dt_long$treatment)))

  # the line follows the group mean at each timepoint, the dots are the
  # individual replicates sitting behind that mean
  dt_mean <- dt_long %>%
    group_by(Genes, treatment, time) %>%
    summarise(Expression = mean(Expression), .groups = "drop")

  # same dodge on both layers, otherwise the dots drift away from their line
  pd <- position_dodge(width = 0.5)

  # name each line at its own last timepoint, so the treatment can be read off
  # the plot without going to the legend. Treatments end at different times
  # (naive at 0h, NC at 72h, cytokines at 120h), which spreads the labels out.
  dt_label <- dt_mean %>%
    group_by(Genes, treatment) %>%
    filter(as.integer(time) == max(as.integer(time))) %>%
    ungroup()

  ggplot(dt_long, aes(x = time, y = Expression, colour = treatment, group = treatment)) +
    geom_line(data = dt_mean, linewidth = 0.7, position = pd) +
    geom_point(size = 1.4, alpha = 0.8, position = pd) +
    ggrepel::geom_text_repel(data = dt_label, aes(label = treatment),
                             size = 2.5, hjust = 0, nudge_x = 0.15,
                             direction = "y", segment.size = 0.2,
                             min.segment.length = 0, show.legend = FALSE) +
    facet_wrap(~Genes, ncol = ncol, scales = "free_y") +
    scale_colour_manual(values = treatment_colours) +
    # extra room on the right so the labels are not clipped
    scale_x_discrete(guide = guide_axis(angle = 90),
                     expand = expansion(mult = c(0.05, 0.22))) +
    labs(title = title, y = "log2(CPM + 1)", x = NULL, colour = "Treatment") +
    theme_bw()
}

# the genes that are most significant anywhere within one set, ranked by their
# best FDR across that set's contrasts
top_genes_for_set <- function(top_tables, n = 6){
  imap_dfr(top_tables, function(tbl, nm) tbl %>% dplyr::select(gene, FDR)) %>%
    group_by(gene) %>%
    summarise(best_FDR = min(FDR), .groups = "drop") %>%
    arrange(best_FDR) %>%
    head(n) %>%
    pull(gene)
}

system(paste("mkdir -p", paste0(imageDir,"genes")))
```

##### 3.5.1 Fixed marker list

```{r plot_counts, fig.width = 10, fig.height = 8, warning=FALSE}
#Select genes to plot
genes_to_plot <- c("Batf", "Nr4a1", "Id3", "Gzmb", "Tnf", "Ifng", "Tcf7")

p <- plot_genes(genes_to_plot, "Marker genes")
p

ggsave(paste0(imageDir,"genes/marker_genes.pdf"), p, width = 10, height = 8)
```

##### 3.5.2 Top genes for each set

```{r plot_counts_per_set, fig.width = 10, fig.height = 8, warning=FALSE, results = "asis"}
# Six most significant genes within each set. Every set gets its own view,
# because the genes driving "IL18+IL36b vs IL2" are not the genes driving
# "everything vs naive".
n_top_genes <- 6

for(set_name in names(top_tables_sets)){
  cat("\n\n###### Set ", set_name, "\n\n", sep="")
  genes <- top_genes_for_set(top_tables_sets[[set_name]], n = n_top_genes)
  cat("Top genes: ", paste(genes, collapse = ", "), "\n\n", sep="")

  p <- plot_genes(genes, paste("Top", n_top_genes, "genes, set", set_name))
  if(is.null(p)){
    cat("No genes to plot.\n\n")
  } else {
    print(p)
    ggsave(paste0(imageDir,"genes/top_genes_", set_files[[set_name]], ".pdf"),
           p, width = 10, height = 8)
    cat("\n\n")
  }
}
```

#### 3.6 MHC class II module

```{r mhc2_module, fig.width = 12, fig.height = 11, warning = FALSE}
# Mouse symbols, not the human HLA-D names: H2-Aa / H2-Ab1 are I-A, H2-Eb1 is
# I-E, H2-DMa / H2-DMb1 / H2-DMb2 are DM, H2-Oa / H2-Ob are DO. There is no
# mouse orthologue of HLA-DP. H2-Ea is absent from this matrix, which fits a
# C57BL/6 (H2b) background where Ea is a pseudogene.
#
# Cd74 is the invariant chain, Ciita the class II transactivator - the
# inducible master switch - and Rfx5 / Rfxank / Rfxap the RFX complex, which is
# constitutively expressed rather than induced. Nlrc5 is the class I
# transactivator, included as a specificity control: if it moves as much as
# Ciita then this is general antigen-presentation upregulation rather than a
# class II specific programme.
mhc2_module <- c("Cd74","Ciita","Rfx5","Rfxank","Rfxap",
                 "H2-Aa","H2-Ab1","H2-Eb1",
                 "H2-DMa","H2-DMb1","H2-DMb2",
                 "H2-Oa","H2-Ob","Nlrc5")

p <- plot_genes(mhc2_module, "MHC class II module over the timecourse", ncol = 4)
p

ggsave(paste0(imageDir,"genes/MHC_class_II_module.pdf"), p, width = 12, height = 11)
```

```{r mhc2_core, fig.width = 10, fig.height = 6, warning = FALSE}
# the core of the module on its own, which is where the late induction sits
p <- plot_genes(c("Cd74","Ciita","H2-Aa","H2-Eb1","Rfx5","Nlrc5"),
                "MHC class II core: invariant chain, transactivator, class II genes",
                ncol = 3)
p

ggsave(paste0(imageDir,"genes/MHC_class_II_core.pdf"), p, width = 10, height = 6)
```

#### 3.7 Export values for Prism

```{r prism_export, warning = FALSE}
# Expression values written out for hand-plotting in Prism. Three scales,
# because they answer different questions:
#   raw_count - featureCounts output after UMI deduplication, still carries
#               library size differences
#   cpm       - TMM-normalised counts per million, depth corrected
#   log2cpm   - log2(CPM + 1), the scale every figure in this document uses
# Taken from the annotated matrix before filterByExpr, so a lowly expressed
# gene is still exported rather than silently missing.
prism_genes <- c("Batf","Id3","Tcf7","Nr4a1","Bach2","Bcl6","Ezh2",
                 # MHC class II module
                 "Cd74","Ciita","Rfx5","Rfxank","Rfxap",
                 "H2-Aa","H2-Ab1","H2-Eb1","H2-DMa","H2-DMb1","H2-DMb2",
                 "H2-Oa","H2-Ob","Nlrc5")

prism_dir <- paste0(tableDir,"prism/")
system(paste("mkdir -p", prism_dir))

# resolve case-insensitively, and say what is missing rather than failing
found <- prism_genes[tolower(prism_genes) %in% tolower(rownames(counts))]
missing_genes <- setdiff(prism_genes, found)
if(length(missing_genes) > 0){
  cat("NOT FOUND in the matrix:", paste(missing_genes, collapse = ", "), "\n")
}
idx <- match(tolower(found), tolower(rownames(counts)))

cpm_mat  <- cpm(dge, log = FALSE)
lcpm_mat <- log2(cpm_mat + 1)

# ---- long format: one row per gene x sample, all metadata attached ----
long <- do.call(rbind, lapply(seq_along(found), function(i){
  r <- idx[i]
  data.frame(
    gene      = found[i],
    sample    = sample_key$sample_name,
    label     = sample_key$label,
    treatment = as.character(sample_key$treatment),
    time      = as.character(sample_key$time),
    replicate = sample_key$replicate,
    batch     = sample_key$batch,
    raw_count = as.numeric(counts[r, ]),
    cpm       = round(as.numeric(cpm_mat[r, ]), 3),
    log2cpm   = round(as.numeric(lcpm_mat[r, ]), 4),
    stringsAsFactors = FALSE
  )
}))

write.csv(long, paste0(prism_dir,"gene_values_long.csv"), row.names = FALSE)
```

```{r prism_export_wide, warning = FALSE}
# ---- wide format, one file per gene per scale ----
# Rows are timepoints, columns are treatment_repN. That is the shape a Prism
# Grouped table expects: paste it in, then group the replicate columns.
# Empty cells are real gaps in the design (naive exists only at 0h, NC stops
# at 72h, IL715_24h has two replicates), so they are written as blanks.
time_levels <- levels(design_matrix$time)
trt_levels  <- levels(design_matrix$treatment)

make_wide <- function(dat, value_col){
  dat$col <- paste0(dat$treatment, "_rep", dat$replicate)

  col_order <- unlist(lapply(trt_levels, function(t){
    reps <- sort(unique(dat$replicate[dat$treatment == t]))
    if(length(reps) == 0) return(NULL)
    paste0(t, "_rep", reps)
  }))

  w <- dat %>%
    dplyr::select(time, col, value = dplyr::all_of(value_col)) %>%
    pivot_wider(names_from = col, values_from = value) %>%
    mutate(time = factor(time, levels = time_levels)) %>%
    arrange(time)

  w[, c("time", intersect(col_order, colnames(w))), drop = FALSE]
}

for(g in found){
  d <- long %>% filter(gene == g)
  write.csv(make_wide(d, "log2cpm"),
            paste0(prism_dir, g, "_log2cpm_wide.csv"), row.names = FALSE, na = "")
  write.csv(make_wide(d, "raw_count"),
            paste0(prism_dir, g, "_rawcount_wide.csv"), row.names = FALSE, na = "")
}

cat("genes exported:", paste(found, collapse = ", "), "\n")
cat("files written to:", prism_dir, "\n\n")
print(list.files(prism_dir))
```

### 4. Pathway analysis

```{r pathway_functions, include = FALSE}
#get the hallmarks database
h_gene_sets = msigdbr(species = "mouse", category = "H")
local_gene_sets_df <- h_gene_sets %>%
  dplyr::select(gs_name, gene_symbol)

# run enrichment separately on the up and down gene lists of each contrast,
# returning NULL when a list is too short to test
run_enrichment <- function(genes, direction_label, treatment_label, min_size = 5){
  genes <- unique(na.omit(genes))
  if(length(genes) < min_size) return(NULL)

  enr <- clusterProfiler::enricher(gene = genes,
                                   TERM2GENE = local_gene_sets_df,
                                   pAdjustMethod = "BH")
  if(is.null(enr)) return(NULL)
  res <- as.data.frame(enr)
  if(nrow(res) == 0) return(NULL)

  res %>%
    mutate(direction = direction_label, treatment = treatment_label)
}

# runs the above across every contrast of one set and collates the result
enrich_set <- function(top_tables, combinations){
  results <- list()
  for(combination in names(top_tables)){
    tbl <- top_tables[[combination]]
    up   <- tbl %>% filter(FDR < 0.05, logFC >= 1)  %>% pull(gene)
    down <- tbl %>% filter(FDR < 0.05, logFC <= -1) %>% pull(gene)

    results[[paste0(combination,"_up")]]   <- run_enrichment(up,   "Upregulated",   combination)
    results[[paste0(combination,"_down")]] <- run_enrichment(down, "Downregulated", combination)
  }

  collated <- bind_rows(results)
  if(nrow(collated) == 0) return(collated)

  collated %>%
    mutate(pathway = gsub("HALLMARK_", "", ID)) %>%
    mutate(
      treatment = forcats::fct_relevel(treatment, names(combinations)),
      pathway = fct_rev(fct_relevel(pathway, sort(unique(pathway))))
    ) %>%
    filter(p.adjust < 0.05)
}

plot_enrichment <- function(collated, title){
  if(nrow(collated) == 0) return(NULL)
  collated %>%
    ggplot(aes(treatment, pathway, size = FoldEnrichment)) +
    geom_point(aes(color = direction)) +
    facet_wrap(~direction, nrow = 1) +
    scale_color_manual(values = c("Upregulated" = "firebrick",
                                  "Downregulated" = "steelblue")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
          strip.text = element_text(size = 8, face = "bold"),
          legend.position = "right") +
    labs(title = title, x = "Treatment", y = "Pathway")
}

show_enrichment_table <- function(collated){
  if(nrow(collated) == 0) return(invisible(NULL))
  datatable(
    collated %>%
      dplyr::select(treatment, direction, pathway, GeneRatio, BgRatio, p.adjust, Count) %>%
      mutate(p.adjust = signif(p.adjust, 3)) %>%
      arrange(p.adjust),
    extensions = 'Buttons',
    options = list(pageLength = 15, scrollX = TRUE, dom = 'Bfrtip',
                   buttons = c('copy','csv','excel')))
}
```

```{r pathway_run, include = FALSE}
# enrichment for all four sets, kept in one named list
collated_sets <- list()
for(set_name in names(top_tables_sets)){
  collated_sets[[set_name]] <- enrich_set(top_tables_sets[[set_name]],
                                          combinations_sets[[set_name]])
  write.csv(collated_sets[[set_name]],
            paste0(tableDir,"hallmark_enrichment_", set_files[[set_name]], ".csv"),
            row.names = FALSE)
}
```

```{r pathway_plots, fig.width = 12, fig.height = 11, warning=FALSE, results = "asis"}
for(set_name in names(collated_sets)){
  cat("\n\n#### Set ", set_name, "\n\n", sep="")
  p <- plot_enrichment(collated_sets[[set_name]], paste("Hallmark enrichment, set", set_name))
  if(is.null(p)){
    cat("No hallmark sets reached significance in this set.\n\n")
  } else {
    print(p)
    ggsave(paste0(imageDir,"hallmark_enrichment_", set_files[[set_name]], ".pdf"),
           p, width = 12, height = 11)
    cat("\n\n")
  }
}
```

#### 4.5 Enrichment tables

##### 4.5.1 Set 1

```{r enrich_table_1, echo=FALSE}
show_enrichment_table(collated_sets[["1_vs_naive"]])
```

##### 4.5.2 Set 2

```{r enrich_table_2, echo=FALSE}
show_enrichment_table(collated_sets[["2_vs_NC"]])
```

##### 4.5.3 Set 3

```{r enrich_table_3, echo=FALSE}
show_enrichment_table(collated_sets[["3_IL18_IL36b_vs_IL2"]])
```

##### 4.5.4 Set 4

```{r enrich_table_4, echo=FALSE}
show_enrichment_table(collated_sets[["4_IL18_IL36b_vs_IL2IL715"]])
```

### 5. Transcription factor activity (DoRothEA)

```{r dorothea_setup, warning = FALSE}
library(decoupleR)
library(dorothea)

# DoRothEA mouse regulons at confidence A-C. This data ships inside the package
# so it works offline; CollecTRI via decoupleR::get_collectri() is the newer
# alternative but needs a live call out to OmnipathR.
data(dorothea_mm, package = "dorothea")

regulons <- dorothea_mm %>%
  filter(confidence %in% c("A","B","C")) %>%
  dplyr::select(source = tf, target, mor)

cat("TFs:", length(unique(regulons$source)), " edges:", nrow(regulons), "\n")
```

```{r tf_functions, include = FALSE}
# One signed statistic per gene per contrast. edgeR's quasi-likelihood test
# reports F, which has no direction, so it is signed by the fold change:
# sign(logFC) * sqrt(F) is monotone in the p-value and keeps direction. Using
# this rather than logFC alone means a large but noisy fold change does not
# outweigh a smaller, well-supported one.
make_stat_matrix <- function(top_tables){
  m <- imap_dfr(top_tables, function(tbl, nm){
    tbl %>% transmute(gene, contrast = nm, stat = sign(logFC) * sqrt(.data[["F"]]))
  }) %>%
    pivot_wider(names_from = contrast, values_from = stat, values_fn = dplyr::first) %>%
    tibble::column_to_rownames("gene") %>%
    as.matrix()
  # a gene missing from any contrast would propagate NAs through the model
  m[complete.cases(m), , drop = FALSE]
}

run_tf_activity <- function(top_tables){
  decoupleR::run_ulm(mat = make_stat_matrix(top_tables),
                     network = regulons, minsize = 5)
}

# The question "which TFs are most influencing this timepoint" answered
# directly: one row per contrast, columns are ranks 1..n, cells are the TF and
# its activity score. Rendered with kable rather than a DT widget because four
# sets of DT tables would add several MB to the document for no extra use.
top_tf_table <- function(tf_acts, contrast_order, n = 10){
  tf_acts %>%
    mutate(condition = factor(condition, levels = contrast_order)) %>%
    group_by(condition) %>%
    slice_max(abs(score), n = n, with_ties = FALSE) %>%
    mutate(rank  = row_number(),
           entry = paste0(source, " (", round(score, 1), ")")) %>%
    ungroup() %>%
    dplyr::select(condition, rank, entry) %>%
    pivot_wider(names_from = rank, values_from = entry, names_prefix = "") %>%
    arrange(condition) %>%
    dplyr::rename(contrast = condition)
}

tf_heatmap <- function(tf_acts, contrast_order, title, file, n_tf = 30){
  top_tfs <- tf_acts %>%
    group_by(source) %>%
    summarise(max_abs = max(abs(score)), .groups = "drop") %>%
    arrange(desc(max_abs)) %>% head(n_tf) %>% pull(source)

  m <- tf_acts %>%
    filter(source %in% top_tfs) %>%
    dplyr::select(source, condition, score) %>%
    pivot_wider(names_from = condition, values_from = score) %>%
    tibble::column_to_rownames("source") %>%
    as.matrix()

  # keep contrasts in the order the set defines them, not alphabetical
  m <- m[, intersect(contrast_order, colnames(m)), drop = FALSE]

  # diverging scale centred on zero: blue = less active, red = more active
  lim <- max(abs(m), na.rm = TRUE)
  br  <- seq(-lim, lim, length.out = 101)
  col <- colorRampPalette(c("#0096FF","white","#F20D0D"))(100)

  pheatmap(m, color = col, breaks = br, cluster_cols = FALSE, cluster_rows = TRUE,
           fontsize_row = 8, fontsize_col = 8, main = title,
           filename = paste0(imageDir, file), width = 12, height = 10)
  pheatmap(m, color = col, breaks = br, cluster_cols = FALSE, cluster_rows = TRUE,
           fontsize_row = 8, fontsize_col = 8, main = title)
}
```

```{r tf_run_all, include = FALSE}
# every set gets its own TF activity, written out in full to CSV
tf_acts_sets <- list()
for(set_name in names(top_tables_sets)){
  tf_acts_sets[[set_name]] <- run_tf_activity(top_tables_sets[[set_name]])
  write.csv(tf_acts_sets[[set_name]],
            paste0(tableDir,"TF_activity_dorothea_", set_files[[set_name]], ".csv"),
            row.names = FALSE)
}
```

#### 5.1 Top TFs per contrast

```{r tf_tables, results = "asis"}
# one table per set: for each contrast, the ten TFs with the largest absolute
# activity score, in rank order
for(set_name in names(tf_acts_sets)){
  cat("\n\n##### Set ", set_name, "\n\n", sep="")
  tab <- top_tf_table(tf_acts_sets[[set_name]], names(top_tables_sets[[set_name]]), n = 10)
  print(knitr::kable(tab, format = "html", escape = FALSE))
  cat("\n\n")
}
```

#### 5.2 Activity heatmaps

```{r tf_heatmaps, fig.width = 12, fig.height = 10, results = "asis"}
for(set_name in names(tf_acts_sets)){
  cat("\n\n##### Set ", set_name, "\n\n", sep="")
  tf_heatmap(tf_acts_sets[[set_name]],
             names(top_tables_sets[[set_name]]),
             paste("DoRothEA TF activity, set", set_name),
             paste0("TF_activity_dorothea_", set_files[[set_name]], ".pdf"))
  cat("\n\n")
}
```

#### 5.3 Batf across all sets

```{r tf_batf, fig.width = 12, fig.height = 8}
# Batf specifically. Activity here is inferred from Batf's targets, so it is a
# different quantity from Batf's own expression - the two are worth comparing
# against the line plots in section 3.5.
batf_all <- imap_dfr(tf_acts_sets, function(x, set_name){
  x %>% filter(tolower(source) == "batf") %>% mutate(set = set_name)
})

if(nrow(batf_all) == 0){
  cat("Batf is not present in the DoRothEA A-C regulons.\n")
} else {
  batf_all <- batf_all %>%
    mutate(condition = factor(condition, levels = unlist(lapply(top_tables_sets, names))))

  p <- ggplot(batf_all, aes(x = condition, y = score, fill = score > 0)) +
    geom_col(colour = "black", width = 0.7) +
    geom_hline(yintercept = 0) +
    facet_wrap(~set, scales = "free_x", nrow = 2) +
    scale_fill_manual(values = c("TRUE" = "#F20D0D", "FALSE" = "#0096FF"), guide = "none") +
    labs(title = "Batf inferred activity across all four contrast sets",
         subtitle = "ULM score from DoRothEA A-C targets",
         x = NULL, y = "Activity score") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7))
  p
  ggsave(paste0(imageDir,"TF_activity_Batf_all_sets.pdf"), p, width = 12, height = 8)
}
```

#### 5.4 Differential programmes

```{r tf_programmes, fig.width = 11, fig.height = 15, results = "asis"}
# TFs grouped into the programmes they belong to, for the two cytokine-versus-
# cytokine sets. Both sides are time-matched here, so a score is a statement
# about the cytokines rather than about time in culture.
#
# SIGN CONVENTION: positive (red) = higher in the IL18 + IL36b arm,
#                  negative (blue) = higher in the IL2 / IL7-15 arm.
# MAPK has no TF of its own - it acts through the transcription factors that
# ERK, JNK and p38 phosphorylate. Two blocks: the AP-1 dimers (JNK/ERK), and
# the ternary-complex / immediate-early / CREB arm (ERK / p38). Cebpb is a p38
# target and sits with the latter. Batf appears here AND under T cell fate; the
# join below gives it a row in each facet.
programmes <- list(
  "NF-kB"           = c("Rela","Rel","Nfkb1","Nfkb2","Relb"),
  "MAPK: AP-1"      = c("Jun","Junb","Jund","Fos","Fosl1","Fosl2","Atf2","Atf3","Batf"),
  "MAPK: ETS/EGR/CREB" = c("Elk1","Ets1","Ets2","Srf","Egr1","Creb1","Atf4","Cebpb"),
  "JAK-STAT"        = c("Stat1","Stat3","Stat4","Stat5a","Stat5b","Stat6"),
  "Proliferation"   = c("Myc","Mycn","E2f1","E2f4","Foxm1","Tfdp1"),
  "MHC class II"    = c("Rfx5","Rfxank","Rfxap","Ciita"),
  "T cell fate"     = c("Tbx21","Irf4","Batf","Prdm1","Tcf7")
)

prog_key <- data.frame(
  source    = unlist(programmes, use.names = FALSE),
  programme = rep(names(programmes), lengths(programmes)),
  stringsAsFactors = FALSE
)

programme_sets <- c("3_IL18_IL36b_vs_IL2", "4_IL18_IL36b_vs_IL2IL715")

for(set_name in programme_sets){
  cat("\n\n##### Set ", set_name, "\n\n", sep="")

  d <- tf_acts_sets[[set_name]] %>%
    filter(source %in% prog_key$source) %>%
    # sets 3 and 4 name their contrasts ..._<timepoint>, so the timepoint can
    # be taken off the end. Sets 1 and 2 do not, which is why they are excluded.
    mutate(tp = factor(sub("^.*_([0-9]+h)$", "\\1", condition),
                       levels = timepoints_all)) %>%
    left_join(prog_key, by = "source") %>%
    mutate(programme = factor(programme, levels = names(programmes)))

  lim <- max(abs(d$score), na.rm = TRUE)

  p <- ggplot(d, aes(x = tp, y = source, fill = score)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = round(score, 1)), size = 2.6, colour = "grey15") +
    facet_grid(programme ~ ., scales = "free_y", space = "free_y") +
    scale_fill_gradient2(low = "#0096FF", mid = "white", high = "#F20D0D",
                         midpoint = 0, limits = c(-lim, lim)) +
    labs(title = paste("Differential programmes, set", set_name),
         subtitle = "Red = higher in IL18 + IL36b; blue = higher in IL2 / IL7-15",
         x = NULL, y = NULL, fill = "ULM score") +
    theme_minimal(base_size = 11) +
    theme(panel.grid = element_blank(),
          strip.text.y = element_text(angle = 0, face = "bold"))
  print(p)

  ggsave(paste0(imageDir,"TF_programmes_", set_files[[set_name]], ".pdf"),
         p, width = 11, height = 10)
  cat("\n\n")
}
```

#### 5.5 Per-cytokine TF activity trajectories

```{r tf_sample_activity, include = FALSE}
# Everything above scores CONTRASTS. This scores SAMPLES, which is what lets
# each treatment - NC included - have a trajectory of its own rather than only
# appearing as the thing being subtracted.
#
# Genes are z-scaled across samples first. Without that, the score is dominated
# by how highly a TF's targets are expressed in absolute terms, which is the
# same in every sample and tells you nothing. After scaling, a score says how
# active a TF looks in that sample relative to the average sample, so 0 means
# average rather than "no change versus a control".
mat_z <- count_matrix[apply(count_matrix, 1, sd) > 0, , drop = FALSE]
mat_z <- t(scale(t(mat_z)))

tf_sample <- decoupleR::run_ulm(mat = mat_z, network = regulons, minsize = 5) %>%
  left_join(sample_key %>% dplyr::select(column, treatment, time, replicate, batch),
            by = c("condition" = "column"))

stopifnot(!any(is.na(tf_sample$treatment)))

write.csv(tf_sample, paste0(tableDir,"TF_activity_per_sample.csv"), row.names = FALSE)
cat("TFs scored per sample:", length(unique(tf_sample$source)),
    " samples:", length(unique(tf_sample$condition)), "\n")
```

```{r tf_trajectories, fig.width = 13, fig.height = 8}
# NF-kB arm (Rela, Rel), Cebpb, the gamma-c STAT arm (Stat5a, Stat5b), Myc,
# and Rfx5. One panel per TF, one line per treatment, so convergence between
# cytokines is visible directly: if IL18 and IL36b track each other on Rela and
# Cebpb while IL2 and IL715 track each other on Stat5 and Myc, that is two
# distinct states rather than one axis of activation strength.
tf_traj <- c("Rela","Rel","Cebpb","Stat5a","Stat5b","Myc","Rfx5")

d <- tf_sample %>%
  filter(source %in% tf_traj) %>%
  mutate(source = factor(source, levels = tf_traj))

d_mean <- d %>%
  group_by(source, treatment, time) %>%
  summarise(score = mean(score), .groups = "drop")

# name each line at its own final timepoint
d_lab <- d_mean %>%
  group_by(source, treatment) %>%
  filter(as.integer(time) == max(as.integer(time))) %>%
  ungroup()

pd <- position_dodge(width = 0.5)

p <- ggplot(d, aes(x = time, y = score, colour = treatment, group = treatment)) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed") +
  geom_line(data = d_mean, linewidth = 0.7, position = pd) +
  geom_point(size = 1.3, alpha = 0.8, position = pd) +
  ggrepel::geom_text_repel(data = d_lab, aes(label = treatment), size = 2.4,
                           hjust = 0, nudge_x = 0.15, direction = "y",
                           segment.size = 0.2, min.segment.length = 0,
                           show.legend = FALSE) +
  facet_wrap(~source, ncol = 4, scales = "free_y") +
  scale_colour_manual(values = treatment_colours) +
  scale_x_discrete(guide = guide_axis(angle = 90),
                   expand = expansion(mult = c(0.05, 0.25))) +
  labs(title = "TF activity trajectories, one line per treatment",
       subtitle = "Per-sample ULM score on gene-wise z-scaled log2(CPM+1). 0 = average sample, dashed line",
       x = NULL, y = "TF activity (ULM score)", colour = "Treatment") +
  theme_bw(base_size = 11)
p

ggsave(paste0(imageDir,"TF_trajectories_per_cytokine.pdf"), p, width = 13, height = 8)
```

```{r tf_trajectories_mapk, fig.width = 13, fig.height = 8}
# The MAPK effectors as per-treatment trajectories. Jun / Junb / Fos are the
# AP-1 core and Batf its T-cell partner, so if IL18 and IL36b drive AP-1 the
# four should move together under those two and not under IL2 / IL7-15. Egr1
# and Elk1 report ERK directly; Ets1 and Creb1 are the broader ERK / p38 arm.
tf_traj_mapk <- c("Jun","Junb","Fos","Batf","Egr1","Elk1","Ets1","Creb1")

d <- tf_sample %>%
  filter(source %in% tf_traj_mapk) %>%
  mutate(source = factor(source, levels = tf_traj_mapk))

d_mean <- d %>% group_by(source, treatment, time) %>%
  summarise(score = mean(score), .groups = "drop")
d_lab <- d_mean %>% group_by(source, treatment) %>%
  filter(as.integer(time) == max(as.integer(time))) %>% ungroup()

p <- ggplot(d, aes(x = time, y = score, colour = treatment, group = treatment)) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed") +
  geom_line(data = d_mean, linewidth = 0.7, position = pd) +
  geom_point(size = 1.3, alpha = 0.8, position = pd) +
  ggrepel::geom_text_repel(data = d_lab, aes(label = treatment), size = 2.4,
                           hjust = 0, nudge_x = 0.15, direction = "y",
                           segment.size = 0.2, min.segment.length = 0, show.legend = FALSE) +
  facet_wrap(~source, ncol = 4, scales = "free_y") +
  scale_colour_manual(values = treatment_colours) +
  scale_x_discrete(guide = guide_axis(angle = 90), expand = expansion(mult = c(0.05, 0.25))) +
  labs(title = "MAPK effector TF activity, one line per treatment",
       subtitle = "Per-sample ULM score on gene-wise z-scaled log2(CPM+1). 0 = average sample",
       x = NULL, y = "TF activity (ULM score)", colour = "Treatment") +
  theme_bw(base_size = 11)
p

ggsave(paste0(imageDir,"TF_trajectories_MAPK.pdf"), p, width = 13, height = 8)
```

#### 5.6 The MHC-II versus JAK-STAT dichotomy

```{r signature_scores, warning = FALSE}
# TF activity is inference from a regulon; this is the direct measurement.
# Two signatures, scored per sample:
#
#   MHC class II - curated mouse module. No Hallmark set covers class II
#     specifically, and the human HLA-D names do not exist in mouse, so this is
#     assembled by hand: invariant chain, transactivator and the H2 loci.
#   IL2/STAT5    - HALLMARK_IL2_STAT5_SIGNALING, the standard gamma-c readout,
#     taken from the mouse msigdbr collection already loaded in section 4.
#
# Scoring is the mean gene-wise z-score across the signature, reusing the same
# scaled matrix as the per-sample TF activity. So 0 is the average sample and
# the two signatures are on a common scale, which is what makes them plottable
# against each other.
mhc2_sig  <- c("Cd74","Ciita","H2-Aa","H2-Ab1","H2-Eb1",
               "H2-DMa","H2-DMb1","H2-DMb2","H2-Oa","H2-Ob")
stat5_sig <- h_gene_sets %>%
  filter(gs_name == "HALLMARK_IL2_STAT5_SIGNALING") %>%
  pull(gene_symbol) %>% unique()

score_signature <- function(genes){
  g <- intersect(genes, rownames(mat_z))
  if(length(g) < 3) stop("too few signature genes found")
  colMeans(mat_z[g, , drop = FALSE])
}

cat("MHC class II genes found :", length(intersect(mhc2_sig,  rownames(mat_z))), "of", length(mhc2_sig),  "\n")
cat("IL2/STAT5 genes found    :", length(intersect(stat5_sig, rownames(mat_z))), "of", length(stat5_sig), "\n")

sig_df <- sample_key %>%
  mutate(`MHC class II` = score_signature(mhc2_sig),
         `IL2/STAT5`    = score_signature(stat5_sig))

write.csv(sig_df %>% dplyr::select(sample_name, treatment, time, replicate,
                                   `MHC class II`, `IL2/STAT5`),
          paste0(tableDir,"signature_scores.csv"), row.names = FALSE)
```

```{r signature_figure, fig.width = 14, fig.height = 6.5, warning = FALSE}
# --- left: each signature as a trajectory per treatment ---
sig_long <- sig_df %>%
  pivot_longer(c(`MHC class II`, `IL2/STAT5`),
               names_to = "signature", values_to = "score")

sig_mean <- sig_long %>%
  group_by(signature, treatment, time) %>%
  summarise(score = mean(score), .groups = "drop")

sig_lab <- sig_mean %>%
  group_by(signature, treatment) %>%
  filter(as.integer(time) == max(as.integer(time))) %>% ungroup()

pd <- position_dodge(width = 0.5)

p1 <- ggplot(sig_long, aes(x = time, y = score, colour = treatment, group = treatment)) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed") +
  geom_line(data = sig_mean, linewidth = 0.8, position = pd) +
  geom_point(size = 1.3, alpha = 0.8, position = pd) +
  ggrepel::geom_text_repel(data = sig_lab, aes(label = treatment), size = 2.5,
                           hjust = 0, nudge_x = 0.15, direction = "y",
                           segment.size = 0.2, min.segment.length = 0,
                           show.legend = FALSE) +
  facet_wrap(~signature, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = treatment_colours) +
  scale_x_discrete(guide = guide_axis(angle = 90),
                   expand = expansion(mult = c(0.05, 0.25))) +
  labs(x = NULL, y = "Signature score (mean z)", colour = "Treatment") +
  theme_bw(base_size = 11)

# --- right: the two signatures against each other, as a state space ---
# If the dichotomy is real, the two arms leave the origin along different axes
# rather than along the same one.
sig_wide_mean <- sig_df %>%
  group_by(treatment, time) %>%
  summarise(`MHC class II` = mean(`MHC class II`),
            `IL2/STAT5`    = mean(`IL2/STAT5`), .groups = "drop") %>%
  arrange(treatment, time)

p2 <- ggplot(sig_df, aes(x = `IL2/STAT5`, y = `MHC class II`, colour = treatment)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(size = 2, alpha = 0.7) +
  geom_path(data = sig_wide_mean, aes(group = treatment), linewidth = 0.8,
            arrow = grid::arrow(length = grid::unit(0.16, "cm"), type = "closed")) +
  ggrepel::geom_text_repel(data = sig_wide_mean, aes(label = gsub("h","",time)),
                           size = 2.4, show.legend = FALSE, max.overlaps = Inf) +
  scale_colour_manual(values = treatment_colours) +
  labs(title = NULL, x = "IL2/STAT5 signature", y = "MHC class II signature",
       colour = "Treatment") +
  theme_bw(base_size = 11)

p <- (p1 | p2) +
  patchwork::plot_annotation(
    title = "MHC class II versus IL2/STAT5 programmes",
    subtitle = "Mean gene-wise z-score per sample. Left: each signature over time. Right: the two against each other, arrows follow the timecourse, numbers are hours"
  )
p

ggsave(paste0(imageDir,"signature_MHCII_vs_STAT5.pdf"), p, width = 14, height = 6.5)
```

### 6. Session info

```{r session_info}
sessionInfo()
```
