---
title: "KEGG pathway maps: JUN / NF-kB signalling under IL2, IL7-15, IL18 and IL36"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{ben_legg_kegg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteBuild{true}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

### 1. Workplace setup

```{r setup, warning = FALSE, include = FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(org.Mm.eg.db)
library(clusterProfiler)
library(pathview)
library(KEGGREST)
library(DT)
```

#### 1.1 Directory structure

```{r directory_setup, warning = FALSE}
# 1. make sure the bioinf_scratch drive is mounted
# 2. make sure you are on a network that can reach KEGG - pathview and
#    enrichKEGG both download from rest.kegg.jp and fail silently offline
# 3. session -> set working directory -> scripts

######## directory structure #######
projectDir=paste0("../")
resultsDir=paste0(projectDir,"/results")
imageDir=paste0(resultsDir,"/figures/")
tableDir = paste(resultsDir,"/tables/",sep="")
# KEGG downloads and the rendered maps go here, not into scripts/
keggDir = paste(resultsDir,"/kegg/",sep="")

system(paste("mkdir -p",imageDir))
system(paste("mkdir -p",tableDir))
system(paste("mkdir -p",keggDir))
```

#### 1.2 What this document does

The differential expression comes from `6a.analysis.Rmd`; this document reads
its contrast tables and paints the fold changes onto KEGG maps. Nothing is
re-fitted here.

KEGG enrichment is run for sets 3 and 4 only (the two time-matched
cytokine-versus-cytokine comparisons). The maps are shown once per set:

| Set | Contrast | Maps per pathway | Colour strips per gene |
|---|---|---|---|
| 2 | each cytokine vs its time-matched NC | three, one each at 24h, 48h, 72h | IL2, IL715, IL18, IL36b |
| 3 | IL18 + IL36b vs IL2 | one | 24h, 48h, 72h, 96h, 120h |
| 4 | IL18 + IL36b vs IL2 + IL7-15 | one | 24h, 48h, 72h, 96h, 120h |

Set 2 shows what each cytokine does on its own against untreated culture.
Sets 3 and 4 show the IL18/IL36b programme against the gamma-c cytokines
through time on a single map.

### 2. Pathway selection

```{r pathway_selection, warning = FALSE}
# Chosen by hand from current KEGG, for maps that carry JUN and NF-kB subunits
# and sit on the IL-1-family (IL18, IL36) or gamma-c (IL2, IL7, IL15) axes.
#
# Not chosen from msigdbr: its KEGG sets are the 2011 LEGACY snapshot, which
# predates the NF-kB (2013), TNF (2014) and IL-17 (2017) maps entirely.
#
# The first block carries both JUN and NF-kB. The last two carry neither but
# are where the four cytokines and their receptors actually live, so they are
# kept as the receptor-level context.
kegg_pathways <- c(
  mmu04064 = "NF-kappa B signaling pathway",
  mmu04010 = "MAPK signaling pathway",
  mmu04660 = "T cell receptor signaling pathway",
  mmu04668 = "TNF signaling pathway",
  mmu04621 = "NOD-like receptor signaling pathway",
  mmu04620 = "Toll-like receptor signaling pathway",
  mmu04657 = "IL-17 signaling pathway",
  # receptor-level context, no JUN
  mmu04630 = "JAK-STAT signaling pathway",
  mmu04060 = "Cytokine-cytokine receptor interaction",
  # the ONLY KEGG mouse map that contains Batf (checked with keggLink). KEGG
  # draws AP-1 as a generic JUN/FOS node on the signalling maps and never lists
  # the dimer partners, so Batf is absent from every map above despite being
  # Jun's partner. This map carries Jun, Fos, Nfkb1, Rela, Batf and Batf3
  # together, and is biologically apt: BATF mediates PD-1 inhibitory signalling
  # in T cells.
  mmu05235 = "PD-L1 expression and PD-1 checkpoint pathway"
)

# genes used to check each map really contains what it was chosen for
check_genes <- list(
  JUN      = c("Jun","Junb","Jund"),
  BATF     = c("Batf","Batf3"),
  NFKB     = c("Nfkb1","Nfkb2","Rela","Relb","Rel"),
  IL2_axis = c("Il2","Il2ra","Il2rb","Il2rg"),
  IL7_15   = c("Il7","Il7r","Il15","Il15ra"),
  IL18     = c("Il18","Il18r1","Il18rap"),
  IL36     = c("Il1rl2","Il1rap","Il36a","Il36b","Il36g","Il1f6","Il1f8","Il1f9")
)
```

```{r pathway_verify, warning = FALSE}
# Pull the current gene membership of each map from KEGG and report which of
# the check genes it contains. Wrapped in tryCatch: if KEGG is unreachable
# this table is skipped rather than killing the document.
verify_pathway <- function(id){
  genes <- tryCatch({
    g <- KEGGREST::keggGet(id)[[1]]$GENE
    # GENE alternates KEGG id, "SYMBOL; description"
    sub(";.*$", "", g[seq(2, length(g), by = 2)])
  }, error = function(e) NULL)

  if(is.null(genes)) return(data.frame(pathway = id, status = "KEGG unreachable"))

  out <- sapply(check_genes, function(cg) paste(intersect(cg, genes), collapse = " "))
  data.frame(pathway = id, name = kegg_pathways[[id]], n_genes = length(genes),
             t(out), stringsAsFactors = FALSE, check.names = FALSE)
}

pathway_check <- bind_rows(lapply(names(kegg_pathways), verify_pathway))
knitr::kable(pathway_check)
```

### 3. Load the contrasts

```{r load_contrasts, warning = FALSE}
# Contrast tables written by 6a.analysis.Rmd. Each is gene, logFC, logCPM, F,
# PValue, FDR with gene symbols as the key.
read_contrast <- function(subdir, name){
  f <- paste0(tableDir, subdir, "/", name, ".csv")
  if(!file.exists(f)) stop("contrast table not found, run 6a.analysis.Rmd first: ", f)
  read.csv(f, stringsAsFactors = FALSE) %>% dplyr::select(gene, logFC, FDR)
}

timepoints    <- c("24h","48h","72h","96h","120h")
timepoints_nc <- c("24h","48h","72h")          # NC only exists at these
cytokines     <- c("IL2","IL715","IL18","IL36b")

# set 2: each cytokine vs its time-matched NC. One list per timepoint, each
# holding the four cytokines, because that is how the maps are drawn.
contrasts_set2 <- setNames(lapply(timepoints_nc, function(tp){
  setNames(lapply(cytokines, function(cyt)
    read_contrast("vs_NC", paste0(cyt, "_", tp, "_vs_NC_", tp))), cytokines)
}), timepoints_nc)

# sets 3 and 4 across the timecourse
contrasts_set3 <- setNames(lapply(timepoints, function(tp)
  read_contrast("IL18_IL36b_vs_IL2", paste0("IL18_IL36b_vs_IL2_", tp))), timepoints)

contrasts_set4 <- setNames(lapply(timepoints, function(tp)
  read_contrast("IL18_IL36b_vs_IL2_IL715", paste0("IL18_IL36b_vs_IL2_IL715_", tp))), timepoints)

# one logFC matrix per map: genes x states, joined by symbol so a gene missing
# from one table stays NA rather than shifting rows.
# purrr:: is spelled out because org.Mm.eg.db loads IRanges, whose reduce()
# masks purrr's and errors on a plain list.
logfc_matrix <- function(contrast_list){
  purrr::reduce(purrr::imap(contrast_list, ~ .x %>% dplyr::select(gene, !!.y := logFC)),
                full_join, by = "gene") %>%
    column_to_rownames("gene") %>%
    as.matrix()
}

mat_set2 <- lapply(contrasts_set2, logfc_matrix)   # one matrix per timepoint
mat_set3 <- logfc_matrix(contrasts_set3)
mat_set4 <- logfc_matrix(contrasts_set4)

cat("genes:", nrow(mat_set3),
    " | set 2 maps per pathway:", length(mat_set2),
    " | set 3/4 strips per gene:", ncol(mat_set3), "\n")
```

```{r entrez_mapping, warning = FALSE}
# pathview wants Entrez IDs. Same bitr mapping as the dane_rnaseq analysis,
# done once on the union of genes and then applied to each matrix.
all_genes <- unique(c(rownames(mat_set3), rownames(mat_set4),
                      unlist(lapply(mat_set2, rownames))))

sym2entrez <- bitr(all_genes, fromType = "SYMBOL", toType = "ENTREZID",
                   OrgDb = org.Mm.eg.db) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  distinct(ENTREZID, .keep_all = TRUE)

cat("symbols mapped to Entrez:", nrow(sym2entrez), "of", length(all_genes), "\n")

to_entrez <- function(m){
  keep <- rownames(m) %in% sym2entrez$SYMBOL
  m <- m[keep, , drop = FALSE]
  rownames(m) <- sym2entrez$ENTREZID[match(rownames(m), sym2entrez$SYMBOL)]
  m
}

mat_set2_e <- lapply(mat_set2, to_entrez)
mat_set3_e <- to_entrez(mat_set3)
mat_set4_e <- to_entrez(mat_set4)
```

### 4. KEGG enrichment, sets 3 and 4

```{r enrich_kegg, warning = FALSE}
# Which KEGG pathways are enriched among the DE genes of the two cytokine-
# versus-cytokine sets at each timepoint:
#   set 3: IL18 + IL36b vs IL2
#   set 4: IL18 + IL36b vs IL2 + IL7-15
# Only these two, because both sides are time-matched cytokine groups; sets 1
# and 2 carry the naive or culture baseline and their enrichment would mostly
# report that. Same enrichKEGG call as the dane_rnaseq analysis. Needs KEGG;
# skipped with a message if unreachable.
enrich_one <- function(tbl, label){
  de <- tbl %>% filter(FDR < 0.05, abs(logFC) >= 1) %>% pull(gene)
  ids <- sym2entrez$ENTREZID[match(de, sym2entrez$SYMBOL)]
  ids <- ids[!is.na(ids)]
  if(length(ids) < 10) return(NULL)

  kk <- tryCatch(enrichKEGG(gene = ids, organism = "mmu", keyType = "kegg",
                            pvalueCutoff = 0.05),
                 error = function(e) NULL)
  if(is.null(kk) || nrow(as.data.frame(kk)) == 0) return(NULL)

  as.data.frame(kk) %>%
    transmute(contrast = label, ID, Description = sub(" - Mus musculus.*$", "", Description),
              GeneRatio, p.adjust = signif(p.adjust, 3), Count)
}

# the two per-set lists are joined into one list BEFORE bind_rows. Handed two
# named lists as separate arguments, bind_rows treats each as a data frame
# whose columns are the timepoints and fails on the size mismatch.
enrich_list <- c(
  purrr::imap(contrasts_set3, ~ enrich_one(.x, paste0("set3_", .y))),
  purrr::imap(contrasts_set4, ~ enrich_one(.x, paste0("set4_", .y)))
)

kegg_enrich <- bind_rows(enrich_list) %>%
  mutate(set = sub("_.*$", "", contrast),
         timepoint = factor(sub("^set[34]_", "", contrast), levels = timepoints)) %>%
  dplyr::select(set, timepoint, ID, Description, GeneRatio, p.adjust, Count)

if(nrow(kegg_enrich) == 0){
  cat("No KEGG enrichment returned (KEGG unreachable, or nothing significant).\n")
} else {
  write.csv(kegg_enrich, paste0(tableDir, "kegg_enrichment_sets3_4.csv"), row.names = FALSE)

  cat("--- of the selected pathways, which are enriched, in which set, and when ---\n")
  knitr::kable(kegg_enrich %>% filter(ID %in% names(kegg_pathways)) %>%
                 arrange(ID, set, timepoint))
}
```

```{r enrich_kegg_table, echo = FALSE}
if(nrow(kegg_enrich) > 0){
  datatable(kegg_enrich, extensions = 'Buttons',
            options = list(pageLength = 15, scrollX = TRUE, dom = 'Bfrtip',
                           buttons = c('copy','csv','excel')))
}
```

### 5. Pathway maps

```{r pathview_function, include = FALSE}
# pathview writes into the working directory, so it is run from inside keggDir
# and the finished PNG path is handed back. Downloads are cached in keggDir
# too, so each map is fetched from KEGG once.
#
# limit caps the colour scale at +/- 2 log2FC, as in the dane_rnaseq analysis:
# signalling genes rarely exceed it, and without the cap a handful of extreme
# genes would wash out everything else.
logfc_cap <- 2

run_pathview <- function(gene_mat, pathway_id, suffix){
  owd <- setwd(keggDir); on.exit(setwd(owd))

  multi <- ncol(gene_mat) > 1
  out <- paste0(pathway_id, ".", suffix, if(multi) ".multi", ".png")

  ok <- tryCatch({
    pathview(gene.data   = gene_mat,
             pathway.id  = pathway_id,
             species     = "mmu",
             limit       = list(gene = logfc_cap),
             low         = list(gene = "#0096FF"),
             mid         = list(gene = "white"),
             high        = list(gene = "#F20D0D"),
             out.suffix  = suffix,
             kegg.dir    = ".",
             multi.state = multi,
             same.layer  = TRUE)
    TRUE
  }, error = function(e){ message(pathway_id, ": ", conditionMessage(e)); FALSE })

  if(ok && file.exists(out)) out else NA_character_
}

# draws one map and embeds it, or says why it could not
show_map <- function(gene_mat, pathway_id, suffix){
  png <- run_pathview(gene_mat, pathway_id, suffix)
  if(is.na(png)){
    cat("Map not produced - KEGG unreachable or pathway unavailable.\n\n")
  } else {
    # path relative to scripts/, which is where the document knits from
    cat("![](", file.path(keggDir, png), ")\n\n", sep = "")
  }
}

strip_note <- function(gene_mat){
  cat("Colour strips per gene, left to right: **",
      paste(colnames(gene_mat), collapse = ", "), "**. ",
      "Red = up, blue = down, capped at +/-", logfc_cap, " log2FC.\n\n", sep = "")
}
```

#### 5.1 Set 2: each cytokine vs its time-matched NC

```{r maps_set2, results = "asis"}
# three maps per pathway, one per timepoint at which an NC exists. Strips are
# the four cytokines, so the map reads as "what does each cytokine do to this
# pathway, on its own, at this time".
strip_note(mat_set2_e[[1]])

for(id in names(kegg_pathways)){
  cat("\n\n##### ", id, " ", kegg_pathways[[id]], "\n\n", sep = "")
  for(tp in timepoints_nc){
    cat("\n\n###### ", tp, "\n\n", sep = "")
    show_map(mat_set2_e[[tp]], id, paste0("set2_", tp))
  }
}
```

#### 5.2 Set 3: IL18 + IL36b vs IL2

```{r maps_set3, results = "asis"}
# one map per pathway, strips are the five timepoints
strip_note(mat_set3_e)

for(id in names(kegg_pathways)){
  cat("\n\n##### ", id, " ", kegg_pathways[[id]], "\n\n", sep = "")
  show_map(mat_set3_e, id, "set3_timecourse")
}
```

#### 5.3 Set 4: IL18 + IL36b vs IL2 + IL7-15

```{r maps_set4, results = "asis"}
strip_note(mat_set4_e)

for(id in names(kegg_pathways)){
  cat("\n\n##### ", id, " ", kegg_pathways[[id]], "\n\n", sep = "")
  show_map(mat_set4_e, id, "set4_timecourse")
}
```

### 6. Session info

```{r session_info}
sessionInfo()
```
