---
title: "Reference projection: ProjecTILs atlases and ImmGen (SingleR)"
output:
  rmarkdown::html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 5
vignette: >
  %\VignetteIndexEntry{ben_legg_projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteBuild{true}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

### 1. Workplace setup

```{r setup, warning = FALSE, include = FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(ggnewscale)
library(scales)
library(pheatmap)
library(Seurat)
library(ProjecTILs)
library(SingleR)
library(celldex)
library(SummarizedExperiment)
library(org.Mm.eg.db)
library(DT)
```

#### 1.1 Directory structure

```{r directory_setup, warning = FALSE}
# 1. make sure the bioinf_scratch drive is mounted
# 2. session -> set working directory -> scripts
# 3. the two ProjecTILs atlases live in ~/resources/projectils/ (see 1.2)

######## directory structure #######
projectDir=paste0("../")
resultsDir=paste0(projectDir,"/results")
imageDir=paste0(resultsDir,"/figures/projections/")
robjectsDir = paste(resultsDir,"/Robjects/",sep="")
tableDir = paste(resultsDir,"/tables/",sep="")

system(paste("mkdir -p",imageDir))
system(paste("mkdir -p",robjectsDir))
system(paste("mkdir -p",tableDir))

refDir <- path.expand("~/resources/projectils/")
```

#### 1.2 Purpose and caveats

Each bulk sample is treated as one "cell" and placed against two kinds of
reference, following the PMC225 analysis:

NC is excluded throughout: it is the DE reference, not a state of interest.

1. **ProjecTILs atlases** (single-cell): the TIL atlas
   (`ref_TILAtlas_mouse_v1`, exhaustion-centred CD8) and the acute-infection
   CD8 atlas (`ref_CD8_LCMV_mouse_v2`, naive to effector to memory).
2. **ImmGen** (bulk, sorted populations) via SingleR, curated to peripheral
   T-cell states. ImmGen contains an *in vitro* activation timecourse
   (`CTR, 1H, 5H, 24H, 48H, 96H`) which lines up with this 24-120h design.

**Caveat, carried over from PMC225.** A bulk sample is a mixture, and a mixture
projected onto a single-cell atlas lands at a centroid rather than in any one
state. The ProjecTILs section is therefore exploratory. SingleR is bulk-to-bulk
and its rank correlation is robust across platforms, so the ImmGen calls are
the more defensible of the two. Neither reference contains cytokine-only
stimulation without TCR engagement, so every call is a nearest resemblance,
not ground truth.

> Reference files, if missing:
> `curl -L -o ref_TILAtlas_mouse_v1.rds https://ndownloader.figshare.com/files/41398167`
> `curl -L -o ref_CD8_LCMV_mouse_v2.rds https://ndownloader.figshare.com/files/41703714`

#### 1.3 Colours

```{r colours}
# same treatment colours as 6a.analysis.Rmd
treatment_colours <- c(
  "Naive" = "gray70", "NC" = "gray40",
  "IL2" = "#F20D0D", "IL715" = "#00B050", "IL18" = "#FF8C00", "IL36b" = "#0096FF"
)

# state palette from the PMC225 projections: naive blue, effector orange/red,
# exhausted dark red; anything unrecognised gets a hue
state_palette <- function(states){
  states <- as.character(unique(states))
  base <- c(Naive="#3E5496", NaiveLike="#3E5496", EarlyActiv="#4DAF4A",
            EffectorMemory="#FF7F00", CentralMemory="#F781BF", Effector="#E41A1C",
            SLEC="#FF4D4D", MemoryPrecursor="#66C2A5", MP="#66C2A5",
            Tpex="#984EA3", Tex="#B2182B", Th1="#1B9E77", Tfh="#E7298A", Treg="#D01556")
  key <- sub("^CD[48]_", "", states)
  col <- base[match(key, names(base))]
  miss <- is.na(col); if(any(miss)) col[miss] <- hue_pal()(sum(miss))
  setNames(as.character(col), states)
}
```

### 2. Build the sample-level object

```{r load_counts}
# Counts from 6a's cache, restricted to the 71 samples 6a kept, using its own
# metadata table so the exclusions cannot drift between the two documents.
load(paste0(robjectsDir, "featurecounts_files.Rdata"))   # df: genes x all 76 samples

meta <- read.csv(paste0(tableDir, "sample_metadata.csv"), stringsAsFactors = FALSE) %>%
  mutate(treatment = factor(treatment, levels = names(treatment_colours)),
         time      = factor(time, levels = c("0h","24h","48h","72h","96h","120h")))

# NC is not projected. It is untreated culture - the right reference for DE,
# but not a state of interest here, and it crowds the atlas plots. Dropping it
# also drops the level, so it disappears from legends and facets.
meta <- meta %>% dplyr::filter(treatment != "NC") %>% mutate(treatment = droplevels(treatment))

missing <- setdiff(meta$sample_name, colnames(df))
if(length(missing) > 0) stop("samples in metadata but not in counts: ", paste(missing, collapse=", "))
df <- df[, meta$sample_name]

# Ensembl -> symbol, collapsing duplicate symbols to the higher-count row.
# Same as 6a section 2.3.
egENSEMBL <- toTable(org.Mm.egENSEMBL)
egSYMBOL  <- toTable(org.Mm.egSYMBOL)
ids <- gsub("\\..*", "", row.names(df)); row.names(df) <- ids
entrez <- egENSEMBL$gene_id[match(row.names(df), egENSEMBL$ensembl_id)]
symbol <- egSYMBOL$symbol[match(entrez, egSYMBOL$gene_id)]
o <- order(rowSums(df), decreasing = TRUE)
df <- df[o, ]; symbol <- symbol[o]
keep <- !is.na(symbol) & !duplicated(symbol)
counts <- as.matrix(df[keep, ]); rownames(counts) <- symbol[keep]

colnames(counts) <- meta$label
rownames(meta)   <- meta$label

cat("genes:", nrow(counts), " samples:", ncol(counts), "\n")
```

```{r seurat_object}
# One Seurat "cell" per bulk sample. Log-normalisation only - no scaling or
# variable-gene selection here, both references do their own.
obj <- CreateSeuratObject(counts = counts, meta.data = meta)
obj <- NormalizeData(obj, verbose = FALSE)
obj
```

### 3. ProjecTILs atlases

```{r projectils_functions}
# Project the samples onto one atlas. Cached: projecting takes a minute or two
# per atlas and the result is deterministic. Returns NULL if the reference
# file is absent so the document still knits.
project_ref <- function(reffile, tag){
  qcache  <- paste0(robjectsDir, "projectils_query_", tag, ".rds")
  bgcache <- paste0(robjectsDir, "projectils_ref_bg_", tag, ".rds")

  if(!file.exists(qcache) || !file.exists(bgcache)){
    if(!file.exists(reffile)) return(NULL)
    ref <- load.reference.map(reffile)

    if(!file.exists(qcache)){
      # filter.cells = FALSE: ProjecTILs' default filter drops anything that
      # does not look like a T cell, and a bulk mixture can fail that test
      q <- make.projection(obj, ref = ref, filter.cells = FALSE, ncores = 1)
      q <- cellstate.predict(ref = ref, query = q)
      saveRDS(q, qcache)
    }
    emb <- Embeddings(ref, "umap")
    bg  <- data.frame(UMAP_1 = emb[,1], UMAP_2 = emb[,2], state = factor(ref$functional.cluster))
    set.seed(1); bg <- bg[sample(nrow(bg), min(6000, nrow(bg))), ]
    saveRDS(bg, bgcache)
  }

  q  <- readRDS(qcache); bg <- readRDS(bgcache)
  proj <- as.data.frame(Embeddings(q, "umap")); colnames(proj) <- c("UMAP_1","UMAP_2")
  proj$state <- q$functional.cluster
  proj$conf  <- q$functional.cluster.conf
  proj$label <- colnames(q)
  proj <- proj %>% left_join(meta, by = "label")

  lv <- union(levels(factor(bg$state)), unique(proj$state))
  bg$state   <- factor(bg$state, levels = lv)
  proj$state <- factor(proj$state, levels = lv)
  list(proj = proj, bg = bg, cols = state_palette(lv))
}

# grey reference with state labels; samples drawn on top
# flip_y reverses the y axis so the atlas sits in the orientation used in the
# published ProjecTILs figures. Everything drawn on top - labels, hulls, arrows -
# is in data coordinates, so it flips with the axis and stays correct.
base_umap <- function(p, flip_y = FALSE){
  cent <- p$bg %>% group_by(state) %>%
    summarise(UMAP_1 = median(UMAP_1), UMAP_2 = median(UMAP_2), .groups = "drop")
  g <- ggplot() +
    geom_point(data = p$bg, aes(UMAP_1, UMAP_2), colour = "grey78", size = 0.7, alpha = 0.7) +
    geom_label(data = cent, aes(UMAP_1, UMAP_2, label = state), size = 2.7, fontface = "bold",
               fill = "white", alpha = 0.7, label.size = 0) +
    theme_bw() + theme(panel.grid = element_blank())
  if(flip_y) g <- g + scale_y_reverse()
  g
}

overview_plot <- function(p, title)
  base_umap(p) +
  geom_point(data = p$proj, aes(UMAP_1, UMAP_2), colour = "black", size = 2.4, alpha = 0.85) +
  ggtitle(title)

# samples coloured by treatment, and the mean position of each treatment at
# each timepoint joined in time order, so the trajectory across the atlas is
# visible. Hours printed at each vertex.
trajectory_plot <- function(p, title, flip_y = FALSE){
  mn <- p$proj %>% group_by(treatment, time) %>%
    summarise(UMAP_1 = mean(UMAP_1), UMAP_2 = mean(UMAP_2), .groups = "drop") %>%
    arrange(treatment, time)
  base_umap(p, flip_y) +
    geom_point(data = p$proj, aes(UMAP_1, UMAP_2, fill = treatment),
               shape = 21, size = 2.6, stroke = 0.2, colour = "grey15", alpha = 0.8) +
    geom_path(data = mn, aes(UMAP_1, UMAP_2, colour = treatment, group = treatment),
              linewidth = 0.8, arrow = grid::arrow(length = grid::unit(0.16,"cm"), type = "closed")) +
    geom_text_repel(data = mn, aes(UMAP_1, UMAP_2, label = gsub("h","",time)),
                    size = 2.6, max.overlaps = Inf, show.legend = FALSE) +
    scale_fill_manual(values = treatment_colours, name = "Treatment") +
    scale_colour_manual(values = treatment_colours, guide = "none") +
    ggtitle(title)
}

# Coloured atlas: every reference cell coloured by its state. alpha controls
# how far it recedes - 0.8 to look at the atlas on its own, ~0.3 as an underlay
# with samples drawn over it.
base_umap_coloured <- function(p, flip_y = FALSE, alpha = 0.8, size = 0.7){
  cent <- p$bg %>% group_by(state) %>%
    summarise(UMAP_1 = median(UMAP_1), UMAP_2 = median(UMAP_2), .groups = "drop")
  g <- ggplot() +
    geom_point(data = p$bg, aes(UMAP_1, UMAP_2, colour = state), size = size, alpha = alpha) +
    scale_colour_manual(values = p$cols, name = "Atlas state",
                        guide = guide_legend(override.aes = list(alpha = 1, size = 3))) +
    geom_label(data = cent, aes(UMAP_1, UMAP_2, label = state), size = 2.7, fontface = "bold",
               fill = "white", alpha = 0.75, label.size = 0) +
    theme_bw() + theme(panel.grid = element_blank())
  if(flip_y) g <- g + scale_y_reverse()
  g
}

# the reference alone, coloured, with no samples - to compare the embedding
# in the .rds against the published atlas figure
reference_plot <- function(p, title, flip_y = FALSE)
  base_umap_coloured(p, flip_y, alpha = 0.8) + ggtitle(title)

# coloured atlas as a transparent underlay, samples and trajectories on top.
# ggnewscale::new_scale_colour() opens a second colour scale so the atlas
# states and the treatments do not fight over one legend.
trajectory_plot_coloured <- function(p, title, flip_y = FALSE){
  mn <- p$proj %>% group_by(treatment, time) %>%
    summarise(UMAP_1 = mean(UMAP_1), UMAP_2 = mean(UMAP_2), .groups = "drop") %>%
    arrange(treatment, time)
  base_umap_coloured(p, flip_y, alpha = 0.3, size = 0.6) +
    ggnewscale::new_scale_colour() +
    geom_point(data = p$proj, aes(UMAP_1, UMAP_2, fill = treatment),
               shape = 21, size = 2.8, stroke = 0.4, colour = "black", alpha = 0.9) +
    geom_path(data = mn, aes(UMAP_1, UMAP_2, colour = treatment, group = treatment),
              linewidth = 0.9, arrow = grid::arrow(length = grid::unit(0.16,"cm"), type = "closed")) +
    geom_text_repel(data = mn, aes(UMAP_1, UMAP_2, label = gsub("h","",time)),
                    size = 2.6, max.overlaps = Inf, show.legend = FALSE,
                    bg.color = "white", bg.r = 0.15) +
    scale_fill_manual(values = treatment_colours, name = "Treatment") +
    scale_colour_manual(values = treatment_colours, guide = "none") +
    ggtitle(title)
}

# predicted state composition for each treatment x timepoint (three replicates).
# dplyr::count is spelled out: SingleR pulls in matrixStats, whose count()
# masks dplyr's and errors on a data frame.
state_composition <- function(p, title){
  cc <- p$proj %>% dplyr::count(treatment, time, state) %>%
    group_by(treatment, time) %>% mutate(frac = n/sum(n)) %>% ungroup()
  ggplot(cc, aes(time, frac, fill = state)) +
    geom_col() +
    facet_wrap(~treatment, nrow = 1) +
    scale_fill_manual(values = p$cols, name = "state") +
    labs(title = title, x = NULL, y = "fraction of replicates") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
}

state_table <- function(p){
  p$proj %>%
    dplyr::select(label, treatment, time, replicate, state, conf) %>%
    mutate(conf = round(conf, 2)) %>%
    arrange(treatment, time, replicate)
}
```

```{r run_projections}
# cache tags carry _noNC so the earlier all-sample projections are not reused
TIL  <- project_ref(paste0(refDir, "ref_TILAtlas_mouse_v1.rds"), "TIL_noNC")
LCMV <- project_ref(paste0(refDir, "ref_CD8_LCMV_mouse_v2.rds"),  "LCMV_noNC")
have_til  <- !is.null(TIL)
have_lcmv <- !is.null(LCMV)
```

```{r ref_status, results = "asis", echo = FALSE}
if(!have_til)  cat("**TIL atlas not found** in `", refDir, "` - section 3.1 is empty.\n\n", sep = "")
if(!have_lcmv) cat("**LCMV atlas not found** in `", refDir, "` - section 3.2 is empty.\n\n", sep = "")
```

#### 3.1 TIL atlas

```{r til_overview, fig.width = 9, fig.height = 7, eval = have_til}
overview_plot(TIL, "Samples (black) on the TIL atlas")
ggsave(paste0(imageDir, "TIL_overview.pdf"), width = 9, height = 7)
```

```{r til_trajectory, fig.width = 10, fig.height = 7.5, eval = have_til}
trajectory_plot(TIL, "TIL atlas: each treatment through the timecourse")
ggsave(paste0(imageDir, "TIL_trajectory.pdf"), width = 10, height = 7.5)
```

```{r til_reference_coloured, fig.width = 10, fig.height = 7.5, eval = have_til}
# the atlas on its own, coloured by state, nothing projected
reference_plot(TIL, "TIL atlas reference, coloured by state")
ggsave(paste0(imageDir, "TIL_reference_coloured.pdf"), width = 10, height = 7.5)
```

```{r til_trajectory_coloured, fig.width = 11, fig.height = 7.5, eval = have_til}
trajectory_plot_coloured(TIL, "TIL atlas: coloured states underneath, treatments through the timecourse on top")
ggsave(paste0(imageDir, "TIL_trajectory_coloured.pdf"), width = 11, height = 7.5)
```

```{r til_composition, fig.width = 13, fig.height = 4.5, eval = have_til}
state_composition(TIL, "TIL atlas: predicted state per treatment and timepoint")
ggsave(paste0(imageDir, "TIL_state_composition.pdf"), width = 13, height = 4.5)
write.csv(state_table(TIL), paste0(tableDir, "projectils_TIL_states.csv"), row.names = FALSE)
```

```{r til_table, echo = FALSE, eval = have_til}
datatable(state_table(TIL), options = list(pageLength = 15))
```

#### 3.2 Acute-infection CD8 atlas (LCMV)

```{r lcmv_overview, fig.width = 9, fig.height = 7, eval = have_lcmv}
overview_plot(LCMV, "Samples (black) on the LCMV CD8 atlas")
ggsave(paste0(imageDir, "LCMV_overview.pdf"), width = 9, height = 7)
```

```{r lcmv_trajectory, fig.width = 10, fig.height = 7.5, eval = have_lcmv}
trajectory_plot(LCMV, "LCMV atlas: each treatment through the timecourse")
ggsave(paste0(imageDir, "LCMV_trajectory.pdf"), width = 10, height = 7.5)
```

```{r lcmv_trajectory_flipped, fig.width = 10, fig.height = 7.5, eval = have_lcmv}
# the same plot with the y axis reversed, matching the orientation of the
# published LCMV atlas so it can sit next to the reference figure in the paper
trajectory_plot(LCMV, "LCMV atlas: each treatment through the timecourse (published orientation, y reversed)",
                flip_y = TRUE)
ggsave(paste0(imageDir, "LCMV_trajectory_flipped.pdf"), width = 10, height = 7.5)
```

```{r lcmv_reference_coloured, fig.width = 10, fig.height = 7.5, eval = have_lcmv}
# the atlas on its own, coloured by state, both orientations
reference_plot(LCMV, "LCMV atlas reference, coloured by state")
ggsave(paste0(imageDir, "LCMV_reference_coloured.pdf"), width = 10, height = 7.5)
reference_plot(LCMV, "LCMV atlas reference, coloured by state (published orientation, y reversed)", flip_y = TRUE)
ggsave(paste0(imageDir, "LCMV_reference_coloured_flipped.pdf"), width = 10, height = 7.5)
```

```{r lcmv_trajectory_coloured, fig.width = 11, fig.height = 7.5, eval = have_lcmv}
trajectory_plot_coloured(LCMV, "LCMV atlas: coloured states underneath, treatments through the timecourse on top")
ggsave(paste0(imageDir, "LCMV_trajectory_coloured.pdf"), width = 11, height = 7.5)
trajectory_plot_coloured(LCMV, "LCMV atlas, coloured underlay (published orientation, y reversed)", flip_y = TRUE)
ggsave(paste0(imageDir, "LCMV_trajectory_coloured_flipped.pdf"), width = 11, height = 7.5)
```

```{r lcmv_composition, fig.width = 13, fig.height = 4.5, eval = have_lcmv}
state_composition(LCMV, "LCMV atlas: predicted state per treatment and timepoint")
ggsave(paste0(imageDir, "LCMV_state_composition.pdf"), width = 13, height = 4.5)
write.csv(state_table(LCMV), paste0(tableDir, "projectils_LCMV_states.csv"), row.names = FALSE)
```

```{r lcmv_table, echo = FALSE, eval = have_lcmv}
datatable(state_table(LCMV), options = list(pageLength = 15))
```

### 4. ImmGen via SingleR

```{r immgen_reference}
# ImmGen curated to peripheral T-cell states, exactly as in PMC225. The
# activation clock (CTR through 96H) is the part that matters most here.
ref <- ImmGenData()
short <- sub(".*\\((.*)\\)", "\\1", ref$label.fine)
state <- dplyr::case_when(
  grepl("^T\\.8Nve|^T\\.8NVE", short)   ~ "CD8 naive",
  grepl("^T\\.4Nve|^T\\.4NVE", short)   ~ "CD4 naive",
  grepl("^T\\.CD8\\.CTR", short)        ~ "CD8 act 0h",
  grepl("^T\\.CD8\\.1H", short)         ~ "CD8 act 1h",
  grepl("^T\\.CD8\\.5H", short)         ~ "CD8 act 5h",
  grepl("^T\\.CD8\\.24H", short)        ~ "CD8 act 24h",
  grepl("^T\\.CD8\\.48H", short)        ~ "CD8 act 48h",
  grepl("^T\\.CD8\\.96H", short)        ~ "CD8 act 96h",
  grepl("^T\\.CD4\\.CTR", short)        ~ "CD4 act 0h",
  grepl("^T\\.CD4\\.1H", short)         ~ "CD4 act 1h",
  grepl("^T\\.CD4\\.5H", short)         ~ "CD4 act 5h",
  grepl("^T\\.CD4\\.24H", short)        ~ "CD4 act 24h",
  grepl("^T\\.CD4\\.48H", short)        ~ "CD4 act 48h",
  grepl("^T\\.CD4\\.96H", short)        ~ "CD4 act 96h",
  grepl("^T\\.8EFF", short)             ~ "CD8 effector",
  grepl("^T\\.8Mem|^T\\.8MEM", short)   ~ "CD8 memory",
  grepl("^T\\.4Mem|^T\\.4MEM", short)   ~ "CD4 memory",
  grepl("Tregs|FP3\\+25\\+", short)     ~ "Treg",
  TRUE ~ NA_character_
)
keep <- !is.na(state)
ref_sub <- ref[, keep]; ref_sub$state <- state[keep]
cat("reference states kept:\n"); print(table(ref_sub$state))

# ordered levels: naive -> activation clock -> effector/memory -> Treg
lv_all <- c("CD8 naive","CD4 naive","CD8 act 0h","CD4 act 0h","CD8 act 1h","CD4 act 1h",
            "CD8 act 5h","CD4 act 5h","CD8 act 24h","CD4 act 24h","CD8 act 48h","CD4 act 48h",
            "CD8 act 96h","CD4 act 96h","CD8 effector","CD8 memory","CD4 memory","Treg")
```

```{r singler_run}
# SingleR on the log-normalised expression, one label per sample. Cached.
cache <- paste0(robjectsDir, "singleR_immgen_noNC.rds")
if(file.exists(cache)){
  pred <- readRDS(cache)
} else {
  qexpr <- GetAssayData(obj, assay = "RNA", layer = "data")
  pred  <- SingleR(test = qexpr, ref = ref_sub, labels = ref_sub$state, de.method = "classic")
  saveRDS(pred, cache)
}

sr <- meta %>%
  mutate(singleR = pred$labels[match(label, rownames(pred))],
         pruned  = pred$pruned.labels[match(label, rownames(pred))])
lv <- lv_all[lv_all %in% sr$singleR]
sr$singleR <- factor(sr$singleR, levels = lv)
scols <- setNames(colorRampPalette(c("#3E5496","#4DAF4A","#FF7F00","#B2182B","#D01556"))(length(lv)), lv)

write.csv(sr %>% dplyr::select(label, treatment, time, replicate, singleR, pruned),
          paste0(tableDir, "singleR_immgen_labels.csv"), row.names = FALSE)

knitr::kable(as.data.frame(table(state = sr$singleR)) %>% filter(Freq > 0) %>% arrange(-Freq))
```

#### 4.1 Label per treatment and timepoint

```{r singler_composition, fig.width = 13, fig.height = 4.5}
cc <- sr %>% dplyr::count(treatment, time, singleR) %>%
  group_by(treatment, time) %>% mutate(frac = n/sum(n)) %>% ungroup()
p <- ggplot(cc, aes(time, frac, fill = singleR)) +
  geom_col() +
  facet_wrap(~treatment, nrow = 1) +
  scale_fill_manual(values = scols, name = "ImmGen state") +
  labs(title = "SingleR / ImmGen label per treatment and timepoint",
       x = NULL, y = "fraction of replicates") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p
ggsave(paste0(imageDir, "immgen_label_composition.pdf"), p, width = 13, height = 4.5)
```

#### 4.2 Similarity to every reference state

```{r singler_scores, fig.width = 12, fig.height = 8}
# The full SingleR score matrix, not just the winning label: how similar each
# sample is to every reference state. Columns ordered treatment then time so
# the activation clock reads left to right within each treatment. This is the
# figure that shows whether the 24-120h culture tracks ImmGen's in-vitro
# activation timecourse, and whether the cytokines differ in where they sit.
sc <- pred$scores
rownames(sc) <- rownames(pred)
sc <- sc[, intersect(lv_all, colnames(sc)), drop = FALSE]

ord <- sr %>% arrange(treatment, time, replicate) %>% pull(label)
sc  <- sc[ord, , drop = FALSE]

# sr carries meta's row names; column_to_rownames refuses a frame that already has them
ann <- sr %>% dplyr::select(label, treatment, time) %>% remove_rownames() %>% column_to_rownames("label")
ann <- ann[ord, , drop = FALSE]
ann_cols <- list(treatment = treatment_colours[levels(droplevels(ann$treatment))],
                 time = setNames(colorRampPalette(c("grey90","grey20"))(nlevels(droplevels(ann$time))),
                                 levels(droplevels(ann$time))))

pheatmap(t(sc),
         cluster_rows = FALSE, cluster_cols = FALSE,
         annotation_col = ann, annotation_colors = ann_cols,
         show_colnames = FALSE,
         color = colorRampPalette(c("white","#FF8C00","#B2182B"))(100),
         main = "SingleR similarity to ImmGen T-cell states",
         filename = paste0(imageDir, "immgen_score_heatmap.pdf"), width = 12, height = 8)
pheatmap(t(sc),
         cluster_rows = FALSE, cluster_cols = FALSE,
         annotation_col = ann, annotation_colors = ann_cols,
         show_colnames = FALSE,
         color = colorRampPalette(c("white","#FF8C00","#B2182B"))(100),
         main = "SingleR similarity to ImmGen T-cell states")
```

#### 4.3 Where on the activation clock

```{r activation_clock, fig.width = 12, fig.height = 5}
# Just the CD8 activation-clock states, as trajectories: for each treatment,
# how similarity to each ImmGen activation timepoint changes across our own
# timecourse. If culture time dominates, all treatments move the same way; if
# a cytokine holds cells earlier or pushes them later, its line separates.
clock <- c("CD8 naive","CD8 act 0h","CD8 act 1h","CD8 act 5h","CD8 act 24h","CD8 act 48h","CD8 act 96h","CD8 effector")
clock <- intersect(clock, colnames(sc))

cl <- as.data.frame(sc[, clock, drop = FALSE]) %>%
  rownames_to_column("label") %>%
  pivot_longer(-label, names_to = "ref_state", values_to = "score") %>%
  left_join(sr %>% dplyr::select(label, treatment, time), by = "label") %>%
  mutate(ref_state = factor(ref_state, levels = clock))

cl_mean <- cl %>% group_by(ref_state, treatment, time) %>%
  summarise(score = mean(score), .groups = "drop")

p <- ggplot(cl_mean, aes(time, score, colour = treatment, group = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_wrap(~ref_state, nrow = 2, scales = "free_y") +
  scale_colour_manual(values = treatment_colours) +
  labs(title = "Similarity to each ImmGen CD8 activation-clock state, over our timecourse",
       x = NULL, y = "SingleR score (Spearman)", colour = "Treatment") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p
ggsave(paste0(imageDir, "immgen_activation_clock.pdf"), p, width = 12, height = 5)
```

### 5. Session info

```{r session_info}
sessionInfo()
```
